---
title: "Chip_seq data analysis"
output: html_document
date: "2025-06-20"
---

This project is to analyse chipseq data for the Degron RPB1 project in collaboration with Scott and Alex from UNSW.

---
Differential signal analysis using extraChips
---

Given that Pol II is a combination of broad and narrow peaks,  unlike histones and TFs which is either broad or narrow, I have decided to go with differential signal analysis which goes by sliding window approach


Tutorial for extraCHIPS : https://bioconductor.org/packages//release/bioc/vignettes/extraChIPs/inst/doc/differential_signal_sliding.html
This part requires more than >50GB of RAM, so this part was deployed on the cluster via R

```{r, run_csaw.R}
#!/usr/bin/env Rscript

library(csaw)
library(extraChIPs)
library(Rsamtools)
library(rtracklayer)
library(glue)
library(here)
library(stringr)
library(magrittr)
library(tidyverse)
library(GenomeInfoDb)

# BAM files
bam_files <- c(
  "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/D8L4Y_bam/D8L4Y_0h_rep1.mLb.clN.sorted.bam",
  "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/D8L4Y_bam/D8L4Y_0h_rep2.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/D8L4Y_bam/D8L4Y_0h_rep3.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/D8L4Y_bam/D8L4Y_90min_rep1.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/D8L4Y_bam/D8L4Y_90min_rep2.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/D8L4Y_bam/D8L4Y_90min_rep3.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/D8L4Y_bam/D8L4Y_8h_rep1.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/D8L4Y_bam/D8L4Y_8h_rep2.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/INP_0h_rep1.mLb.clN.sorted.bam",
  "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/INP_0h_rep2.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/INP_0h_rep3.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/INP_90min_rep1.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/INP_90min_rep2.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/INP_90min_rep3.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/INP_8h_rep1.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/INP_8h_rep2.mLb.clN.sorted.bam"

)

bfl <- BamFileList(bam_files)
names(bfl) <- tools::file_path_sans_ext(basename(bam_files))

# Seqinfo (keeps only the key chromosomes, as the data wats generated from nf-core so no chr prefix)
sq <- seqinfo(BamFile(bam_files[1]))
sq <- keepSeqlevels(sq, c(as.character(1:22), "X", "Y"))

# Grey/black lists
blacklist <- rtracklayer::import.bed(
  "/lmb/home/png/cephfs2/Genomes/merged_hg38_sacCer3/hg38-blacklist.v2.diffsig.bed"
)
blacklist <- keepSeqlevels(blacklist, seqlevels(sq), pruning.mode="coarse")

# Read parameters
rp <- readParam(
  pe = "none",
  dedup = TRUE,
  discard = blacklist
)

# Count windows
wincounts <- windowCounts(
  bam.files = bfl,
  spacing = 40,
  width = 120,
  ext = 200,
  filter = length(bfl),
  param = rp
)


# Save output
#saveRDS(wincounts, file = "Degron_chip_wincounts.rds")
sample <- data.frame(samples = c("D8L4Y_0h_rep1.mLb.clN.sorted",
                                 "D8L4Y_0h_rep2.mLb.clN.sorted",
                                 "D8L4Y_0h_rep3.mLb.clN.sorted",
                                 "D8L4Y_90min_rep1.mLb.clN.sorted",
                                 "D8L4Y_90min_rep2.mLb.clN.sorted",
                                 "D8L4Y_90min_rep3.mLb.clN.sorted",
                                 "D8L4Y_8h_rep1.mLb.clN.sorted",
                                 "D8L4Y_8h_rep2.mLb.clN.sorted",
                                 "INP_0h_rep1.mLb.clN.sorted",
                                 "INP_0h_rep2.mLb.clN.sorted",
                                 "INP_0h_rep3.mLb.clN.sorted",
                                 "INP_90min_rep1.mLb.clN.sorted",
                                 "INP_90min_rep2.mLb.clN.sorted",
                                 "INP_90min_rep3.mLb.clN.sorted",
                                 "INP_8h_rep1.mLb.clN.sorted",
                                 "INP_8h_rep2.mLb.clN.sorted"),
                     group = c("D8L4Y_0h",
                                 "D8L4Y_0h",
                                 "D8L4Y_0h",
                                 "D8L4Y_90min",
                                 "D8L4Y_90min",
                                 "D8L4Y_90min",
                                 "D8L4Y_8h",
                                 "D8L4Y_8h",
                                 "INP_0h",
                                 "INP_0h",
                                 "INP_0h",
                                 "INP_90min",
                                 "INP_90min",
                                 "INP_90min",
                                 "INP_8h",
                                 "INP_8h"))
#add information to the data

colData(wincounts) <- colData(wincounts) %>%
  as_tibble(rownames = "samples") %>% 
  left_join(sample, by = "samples") %>% 
  dplyr::select(
    samples, all_of(colnames(colData(wincounts))), group
  ) %>%
  DataFrame(row.names = .$samples)


  #import consensus narrow peaks file
#peaks <- import("/cephfs2/png/Degron_Scott/Output/chipseq_out/D8L4Y.consensus_peaks.bed")

#replace peaks with genes ie regions of interest
regions <-  import("/cephfs2/png/Genomes/hg38/Homo_sapiens.GRCh38.113_genes_.gtf")
#start the filtering process, match it to the sample's respective input files
samples_list <- colnames(wincounts)

#remove sample3 input for 8h as it is not used for the analysis
samples_list <- samples_list[!samples_list %in% c(
  "INP_8h_rep3.mLb.clN.sorted"
)]

samples_list <- samples_list[!grepl("^INP_", samples_list)]

filtered_list <- list()

for (s in samples_list) {

  idx <- match(s, colnames(wincounts))
  if (is.na(idx)) next

  bg_sample <- sub("^D8L4Y", "INP", s)
  bg_idx <- match(bg_sample, colnames(wincounts))
  if (is.na(bg_idx)) next

  filtered_list[[s]] <- dualFilter(
    x   = wincounts[, idx],
    bg  = wincounts[, bg_idx],
    ref = regions,
    q   = 0.6
  )
  #combined_counts <- do.call(cbind,
  #lapply(filtered_list, assay)
#)
}

saveRDS(filtered_list, "Degron_chip_filtered_counts_regions.RDS")

cat("Finished successfully\n")
```


read in the filtered counts RDS object and proceed with differential signal analysis using sliding window

```{r, diff_signal analysis}
#!/usr/bin/env Rscript

library(csaw)
library(extraChIPs)
library(Rsamtools)
library(rtracklayer)
library(glue)
library(here)
library(stringr)
library(magrittr)
library(tidyverse)
library(GenomeInfoDb)
library(ggplot2)
library(viridis)
library(MASS)
library(ggdag)

filtered_wincounts <- readRDS("Degron_chip_filtered_counts_regions.RDS")

# find common genomic windows
common_ids <- Reduce(intersect,
                     lapply(filtered_wincounts, function(x) {
                         rr <- rowRanges(x)
                         paste0(seqnames(rr), ":", start(rr), "-", end(rr))
                     }))

# subset all objects to same rows
filtered_common <- lapply(filtered_wincounts, function(se) {

    rr <- rowRanges(se)
    ids <- paste0(seqnames(rr), ":", start(rr), "-", end(rr))

    se[ids %in% common_ids, ]
})

# reorder rows identically
filtered_common <- lapply(filtered_common, function(se) {

    rr <- rowRanges(se)
    ids <- paste0(seqnames(rr), ":", start(rr), "-", end(rr))

    se[match(common_ids, ids), ]
})

# combine samples
combined <- do.call(cbind, filtered_common)

#plot assay densities with the common ones
# Assaydensity <- plotAssayDensities(combined,
#                    assay = "logCPM",
#                    colour = "sample") +
#   theme_bw()

#PCA
# PCA <- plotAssayPCA(combined, "logCPM", colour = "group", label = "sample") +
#   theme_bw()
 
#  ggsave(
#     filename = "PCA_extraChips.pdf",
#     plot = PCA,
#     width = 6,
#     height = 4
# )
#RLE
# RLE_plot <-  plotAssayRle(combined, assay = "logCPM", fill = "sample") +
#   geom_hline(yintercept = 0, linetype = 2, colour = "grey") +
#   ggtitle("RLE: Across All Samples") +
#   theme_bw() +
#    theme(axis.text.x = element_text(angle = 90, hjust = 1))

 
#  ggsave(
#     filename = "RLE_extraChips.pdf",
#     plot = RLE_plot,
#     width = 6,
#     height = 4
# )


# Differential region analysis and significance-aware merging
# ____________________________________________________________
#
# This workflow processes differential binding results separately
# for each experimental contrast (90 min vs 0 h and 8 h vs 0 h).
#
# 1. Significance definition:
#    For each fitted model, regions are classified as significant
#    based on an FDR threshold (< 0.05).
#
# 2. Conditional region merging:
#    - If no significant regions are detected, the original
#      window-level results are retained and placeholder values
#      are assigned for merged p-values.
#    - If significant regions exist, adjacent significant windows
#      are merged using a harmonic mean p-value (HMP) approach
#      within a specified genomic distance (merge_within = 120 bp).
#
# 3. Post-processing:
#    - Merged regions are annotated with differential status
#      (via addDiffStatus).
#    - This allows downstream visualization and interpretation
#      of region-level differential signal.
#
# 4. Purpose:
#    This step reduces redundancy from overlapping sliding windows
#    and enables identification of broader genomic regions of
#    consistent differential signal across conditions.
#
# NOTE:
# Each contrast (fit_90, fit_8h) is processed independently to
# preserve condition-specific biological signals.
#
# ============================================================
library(scales)
X <- model.matrix(~group, data = colData(combined))

#X
#                                (Intercept) groupD8L4Y_8h groupD8L4Y_90min
#D8L4Y_0h_rep1.mLb.clN.sorted              1             0                0
#D8L4Y_0h_rep2.mLb.clN.sorted              1             0                0
#D8L4Y_0h_rep3.mLb.clN.sorted              1             0                0
#D8L4Y_90min_rep1.mLb.clN.sorted           1             0                1
#D8L4Y_90min_rep2.mLb.clN.sorted           1             0                1
#D8L4Y_90min_rep3.mLb.clN.sorted           1             0                1
#D8L4Y_8h_rep1.mLb.clN.sorted              1             1                0
#D8L4Y_8h_rep2.mLb.clN.sorted              1             1                0
#attr(,"assign")
#[1] 0 1 1
#attr(,"contrasts")
#attr(,"contrasts")$group
#[1] "contr.treatment"




#differential signal of 90min vs 0h 
fit_90 <- fitAssayDiff(
  combined,
  design = X,
  coef = "groupD8L4Y_90min",
  fc = 1.0, #no foldchange implied, unless i use lfc
  asRanges = TRUE
)

#differential signal of 8h vs 0h
fit_8h <- fitAssayDiff(
  combined,
  design = X,
  coef = "groupD8L4Y_8h",
  fc = 1.0, #no foldchange implied, unless i use lfc
  asRanges = TRUE
)


# all_plot <- fit_gr %>% 
#   as_tibble() %>% 
#   ggplot(aes(logCPM, logFC)) +
#   geom_point(alpha = 0.6) +
#   geom_smooth(se = FALSE, method = "loess") +
#   geom_label(
#     aes(label = label),
#     data = . %>% 
#       summarise(
#         n = dplyr::n(),
#         logCPM = max(logCPM) - 0.15 * diff(range(logCPM)),
#         logFC = max(logFC) - 0.05 * diff(range(logFC)),
#         label = glue("{comma(n)}\nSliding Windows")
#       )
#   ) +
#   ylim(range(fit_gr$logFC)) +
#   ggtitle("MA Plot: All Retained Sliding Windows")



#plot the differential binding result, but did not work because checked there is no signal
fit_90$sig <- fit_90$FDR < 0.05
fit_8h$sig <- fit_8h$FDR < 0.05


if (!any(fit_90$sig, na.rm = TRUE)) {

    results_90_xchip <- fit_90
    mcols(results_90_xchip)$hmPValue <- NA
    mcols(results_90_xchip)$hmPValue_fdr <- NA

} else {

    results_90_xchip <- mergeByHMP(
        fit_90,
        sig_col = "sig",
        merge_within = 120
    ) %>%
        addDiffStatus(
            sig_col = "sig",
            drop = TRUE
        )
}

if (!any(fit_8h$sig, na.rm = TRUE)) {

    results_8h_xchip <- fit_8h
    mcols(results_8h_xchip)$hmPValue <- NA
    mcols(results_8h_xchip)$hmPValue_fdr <- NA

} else {

    results_8h_xchip <- mergeByHMP(
        fit_8h,
        sig_col = "sig",
        merge_within = 120
    ) %>%
        addDiffStatus(
            sig_col = "sig",
            drop = TRUE
        )
}

#check for number of merged_windows which are less than 0.05

table(results_8h_xchip$FDR < 0.05)
table(results_90_xchip$FDR < 0.05)

#cross check the stats before merging using mergeHPM
#> table(results_8h_xchip$FDR < 0.05)
#  FALSE 
# 4013881 
#table(results_90_xchip$FDR < 0.05)
#  FALSE 
# 4013881 

#90min vs 0h
df_90 <- results_90_xchip %>%
  as_tibble() %>%
  mutate(FDR_sig = hmPValue_fdr < 0.05)


#generate the all regions MA plots
FDR_plot_90 <- ggplot(df_90, aes(logCPM, logFC)) +

  geom_point(alpha = 0.5, colour = "grey70") +

  geom_point(
    data = df_90 %>% dplyr::filter(FDR_sig),
    colour = "red",
    alpha = 0.8
  ) +

   geom_hline(yintercept = 0) +
  
  scale_y_continuous(limits = c(-2,2), breaks = c(-2, -1, 0, 1, 2)) +
  annotate(
    "text",
    x = Inf, y = Inf,
    hjust = 1.1, vjust = 1.5,
    label = paste0(
      "FDR < 0.05: ",
      sum(df_90$FDR_sig, na.rm = TRUE),
      " regions\n(out of ",
      nrow(df_90),
      ")"
    )
  ) +
  theme(
    legend.position = "right",
    legend.direction = "vertical",
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    strip.text.x = element_text(size = 12),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 12),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
   panel.background = element_rect(fill = "white", colour = NA)
  ) +
  ggtitle("MA Plot: differential signal analysis 90 min vs 0h") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  remove_grid()

#8h vs 0h
df_8h <- results_8h_xchip %>%
  as_tibble() %>%
  mutate(FDR_sig = hmPValue_fdr < 0.05)


#generate the all regions MA plots
FDR_plot_8h <- ggplot(df_8h, aes(logCPM, logFC)) +

  geom_point(alpha = 0.5, colour = "grey70") +

  geom_point(
    data = df_8h %>% dplyr::filter(FDR_sig),
    colour = "red",
    alpha = 0.8
  ) +

  geom_hline(yintercept = 0) +
scale_y_continuous(limits = c(-2,2),breaks = c(-2, -1, 0, 1, 2)) +
  annotate(
    "text",
    x = Inf, y = Inf,
    hjust = 1.1, vjust = 1.5,
    label = paste0(
      "FDR < 0.05: ",
      sum(df_8h$FDR_sig, na.rm = TRUE),
      " regions\n(out of ",
      nrow(df_8h),
      ")"
    )
  ) +
  theme(
    legend.position = "right",
    legend.direction = "vertical",
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    strip.text.x = element_text(size = 12),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 12),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
   panel.background = element_rect(fill = "white", colour = NA)
  ) +
  ggtitle("MA Plot: differential signal analysis 8h vs 0h") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  remove_grid()

ggsave("MA_xchips_90min_vs_0h.png", FDR_plot_90, width = 8, height = 6, dpi = 300)
ggsave("MA_xchips_8h_vs_0h.png", FDR_plot_8h, width = 8, height = 6, dpi = 300)


#now do by TSS
gtf.dat <- import("/cephfs2/png/Genomes/hg38/Homo_sapiens.GRCh38.113.gtf")
gene.gr <- gtf.dat[gtf.dat$type %in% "gene",]
names(gene.gr) <- gene.gr$gene_id
gene.gr$gene.width <- GenomicRanges::width(gene.gr)
gene.gr$gene_name_column <- gene.gr$gene_name

gene.gr <- gene.gr[width(gene.gr) >1000]
gene.gr <- gene.gr[gene.gr$gene_biotype %in% "protein_coding"]
gene.gr <- gene.gr[countOverlaps(gene.gr,gene.gr) == 1]


#make sure gene.gr is > 1kb and only has non-overlapping protein coding genes
tss <- resize(gene.gr, width = 1, fix = "start")
tss_500 <- promoters(tss, upstream = 30, downstream = 300)
#flank(tss, width = 500, start = TRUE)
tss_hits <- subsetByOverlaps(results_90_xchip, tss_500)

df_tss <- as_tibble(tss_hits) %>%
  mutate(FDR_sig = FDR < 0.05)


n_sig2 <- sum(df_tss$FDR_sig, na.rm = TRUE)
n_tot2 <- nrow(df_tss)


# #plot the TSS only MA density plot
MAplot_TSS_90min <- ggplot(df_tss, aes(x = logCPM, y = logFC)) +

  # optional faint points (same as your res_df2 plot)
  geom_point(alpha = 0.3, size = 0.8, color = "grey") +

  # 2D density (matches your style)
  stat_density_2d(
    aes(fill = after_stat(level), alpha = after_stat(level)),
    geom = "polygon",
    contour = TRUE,
    bins = 20
  ) +

  # FDR significant points
  geom_point(
    data = df_tss %>% dplyr::filter(FDR_sig),
    color = "red",
    size = 0.8
  ) +

  geom_hline(yintercept = 0, linetype = "dotted", color = "black") +

  scale_fill_viridis_c(option = "magma", name = "Density") +
  scale_alpha(range = c(0.2, 0.8), guide = "none") +

  scale_y_continuous(limits = c(-1.5, 1.5),breaks=c(-1.5,-1,0,1,1.5)) +

    theme(
    legend.position = "right",
    legend.direction = "vertical",
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    strip.text.x = element_text(size = 12),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 12),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    panel.background = element_rect(fill = "white", colour = NA)
  )  +
  labs(
    x = "logCPM",
    y = "log2 Fold Change",
    title = "differential signal TSS MA plot with density (90min vs 0h)"
  ) +
  annotate(
  "text",
  x = Inf, y = Inf,
  hjust = 1.1, vjust = 1.5,
  label = paste0(
    "FDR < 0.05: ",
    sum(df_tss$FDR_sig, na.rm = TRUE),
    " / ",
    nrow(df_tss)
  )
) +
  remove_grid()

#8h TSS±500bp
tss_hits_8h <- subsetByOverlaps(results_8h_xchip, tss_500)

df_tss_8h <- as_tibble(tss_hits_8h) %>%
  mutate(FDR_sig = FDR < 0.05)

n_sig <- sum(df_tss_8h$FDR_sig, na.rm = TRUE)
n_tot <- nrow(df_tss_8h)


#generate the TSS desnity MA plot
MAplot_TSS_8h <- ggplot(df_tss_8h, aes(x = logCPM, y = logFC)) +

  # optional faint points (same as your res_df2 plot)
  geom_point(alpha = 0.3, size = 0.8, color = "grey") +

  # 2D density (matches your style)
  stat_density_2d(
    aes(fill = after_stat(level), alpha = after_stat(level)),
    geom = "polygon",
    contour = TRUE,
    bins = 20
  ) +

  # FDR significant points
  geom_point(
    data = df_tss_8h %>% dplyr::filter(FDR_sig),
    color = "red",
    size = 0.8
  ) +

  geom_hline(yintercept = 0, linetype = "dotted", color = "black") +

  scale_fill_viridis_c(option = "magma", name = "Density") +
  scale_alpha(range = c(0.2, 0.8), guide = "none") +

  scale_y_continuous(limits = c(-1.5, 1.5),breaks=c(-1.5,-1,0,1,1.5)) +

    theme(
    legend.position = "right",
    legend.direction = "vertical",
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    strip.text.x = element_text(size = 12),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 12),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    panel.background = element_rect(fill = "white", colour = NA)
  )  +
  labs(
    x = "logCPM",
    y = "log2 Fold Change",
    title = "differential signal TSS MA plot with density (8h vs 0h)"
  ) +
  annotate(
  "text",
  x = Inf, y = Inf,
  hjust = 1.1, vjust = 1.5,
  label = paste0(
    "FDR < 0.05: ",
    sum(df_tss_8h$FDR_sig, na.rm = TRUE),
    " / ",
    nrow(df_tss_8h)
  )
) +
  remove_grid()

ggsave("MA_xchips_TSS_90min_vs_0h.png", MAplot_TSS_90min, width = 8, height = 6, dpi = 300)
ggsave("MA_xchips_TSS_8h_vs_0h.png", MAplot_TSS_8h, width = 8, height = 6, dpi = 300)


###repeat the same for genebody
hg38.gene.body.bed <-  read.table(
    "/cephfs2/png/Genomes/hg38/hg38.gene.body.bed",
    sep = "\t",
    header = TRUE,
    stringsAsFactors = FALSE
)

genebody <- GenomicRanges::makeGRangesFromDataFrame(
  hg38.gene.body.bed ,
  seqnames.field = "chr",
  start.field    = "start",
  end.field      = "end",
  keep.extra.columns = TRUE
)

seqlevels(genebody) <- sub("^chr", "", seqlevels(genebody))
seqnames(genebody) <- sub("^chr", "", seqnames(genebody))


#90mins vs 0h
genebody_hits <- subsetByOverlaps(results_90_xchip, genebody)

df_genebody <- as_tibble(genebody_hits) %>%
  mutate(FDR_sig = FDR < 0.05)


#check the significance statistics
n_sig3 <- sum(df_genebody$FDR_sig, na.rm = TRUE)
n_tot3 <- nrow(df_genebody)


#generate the density MA plot
MAplot_genebody_90mins <- ggplot(df_genebody, aes(x = logCPM, y = logFC)) +

  # optional faint points (same as your res_df2 plot)
  geom_point(alpha = 0.3, size = 0.8, color = "grey") +

  # 2D density (matches your style)
  stat_density_2d(
    aes(fill = after_stat(level), alpha = after_stat(level)),
    geom = "polygon",
    contour = TRUE,
    bins = 20
  ) +

  # FDR significant points
  geom_point(
    data = df_genebody %>% dplyr::filter(FDR_sig),
    color = "red",
    size = 0.8
  ) +

  geom_hline(yintercept = 0, linetype = "dotted", color = "black") +

  scale_fill_viridis_c(option = "magma", name = "Density") +
  scale_alpha(range = c(0.2, 0.8), guide = "none") +

  scale_y_continuous(limits = c(-2, 2)) +

   theme(
    legend.position = "right",
    legend.direction = "vertical",
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    strip.text.x = element_text(size = 12),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 12),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    panel.background = element_rect(fill = "white", colour = NA)
  )  +
  labs(
    x = "logCPM",
    y = "log2 Fold Change",
    title = "differential signal gene body MA plot with density (90min vs 0h)"
  ) +
  annotate(
  "text",
  x = Inf, y = Inf,
  hjust = 1.1, vjust = 1.5,
  label = paste0(
    "FDR < 0.05: ",
    sum(df_genebody$FDR_sig, na.rm = TRUE),
    " / ",
    nrow(df_genebody)
  )
) +
  remove_grid()


#8h vs 0h
genebody_hits_8h <- subsetByOverlaps(results_8h_xchip, genebody)

df_genebody_8h <- as_tibble(genebody_hits_8h) %>%
  mutate(FDR_sig = FDR < 0.05)

#check the significance statistics
n_sig4 <- sum(df_genebody_8h$FDR_sig, na.rm = TRUE)
n_tot4 <- nrow(df_genebody_8h)


#generate the density MA plot
MAplot_genebody_8h <- ggplot(df_genebody_8h, aes(x = logCPM, y = logFC)) +

  # optional faint points (same as your res_df2 plot)
  geom_point(alpha = 0.3, size = 0.8, color = "grey") +

  # 2D density (matches your style)
  stat_density_2d(
    aes(fill = after_stat(level), alpha = after_stat(level)),
    geom = "polygon",
    contour = TRUE,
    bins = 20
  ) +

  # FDR significant points
  geom_point(
    data = df_genebody_8h %>% dplyr::filter(FDR_sig),
    color = "red",
    size = 0.8
  ) +

  geom_hline(yintercept = 0, linetype = "dotted", color = "black") +

  scale_fill_viridis_c(option = "magma", name = "Density") +
  scale_alpha(range = c(0.2, 0.8), guide = "none") +

  scale_y_continuous(limits = c(-2, 2)) +

    theme(
    legend.position = "right",
    legend.direction = "vertical",
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    strip.text.x = element_text(size = 12),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 12),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    panel.background = element_rect(fill = "white", colour = NA)
  )  +
  labs(
    x = "logCPM",
    y = "log2 Fold Change",
    title = "differential signal gene body MA plot with density (8h vs 0h)"
  ) +
  annotate(
  "text",
  x = Inf, y = Inf,
  hjust = 1.1, vjust = 1.5,
  label = paste0(
    "FDR < 0.05: ",
    sum(df_genebody_8h$FDR_sig, na.rm = TRUE),
    " / ",
    nrow(df_genebody_8h)
  )
) +
  remove_grid()

  
ggsave("MA_xchips_genebody_90min_vs_0h.png", MAplot_genebody_90mins, width = 8, height = 6, dpi = 300)
ggsave("MA_xchips_genebody_8h_vs_0h.png", MAplot_genebody_8h, width = 8, height = 6, dpi = 300)

saveRDS(df_tss, "df_tss_90min.rds")
saveRDS(df_tss_8h, "df_tss_8h.rds")

saveRDS(df_genebody, "df_genebody_90min.rds")
saveRDS(df_genebody_8h, "df_genebody_8h.rds")

#move the stats genartion part to local Rstudio
#generating tsv reports
# df_90_out <- results_90_xchip %>%
#   as_tibble() %>%
#   dplyr::select(seqnames, start, end, logCPM, logFC, hmPValue, hmPValue_fdr)

# readr::write_tsv(df_90_out, "xchip_90min_vs_0h_all_regions.tsv")

# df_8h_out <- results_8h_xchip %>%
#   as_tibble() %>%
#   dplyr::select(seqnames, start, end, logCPM, logFC, hmPValue, hmPValue_fdr)

# readr::write_tsv(df_8h_out, "xchip_8h_vs_0h_all_regions.tsv")

# df_tss_out <- df_tss %>%
#   mutate(
#     gene_id = mcols(tss_hits)$gene_id,
#     gene_name = mcols(tss_hits)$gene_name
#   ) %>%
#   dplyr::select(
#     gene_id,
#     gene_name,
#     seqnames,
#     start,
#     end,
#     logCPM,
#     logFC,
#     hmPValue,
#     FDR = hmPValue_fdr,
#     FDR_sig
#   )

# readr::write_tsv(df_tss_out, "xchip_90min_TSS.tsv")

# df_tss_8h_out <- df_tss_8h %>%
#   mutate(
#     gene_id = mcols(tss_hits_8h)$gene_id,
#     gene_name = mcols(tss_hits_8h)$gene_name
#   ) %>%
#   dplyr::select(
#     gene_id,
#     gene_name,
#     seqnames,
#     start,
#     end,
#     logCPM,
#     logFC,
#     hmPValue,
#     FDR = hmPValue_fdr,
#     FDR_sig
#   )

# readr::write_tsv(df_tss_8h_out, "xchip_8h_TSS.tsv")


# df_genebody_90min_out <- df_genebody %>%
#   mutate(
#     gene_id = mcols(genebody_hits)$gene_id,
#     gene_name = mcols(genebody_hits)$gene_name
#   ) %>%
#   dplyr::select(
#     gene_id,
#     gene_name,
#     seqnames,
#     start,
#     end,
#     logCPM,
#     logFC,
#     hmPValue,
#     FDR = hmPValue_fdr,
#     FDR_sig
#   )

# readr::write_tsv(df_genebody_90min_out, "xchip_90min_gene_body.tsv")


# df_genebody_8h_out <- df_genebody_8h %>%
#   mutate(
#     gene_id = mcols(genebody_hits_8h)$gene_id,
#     gene_name = mcols(genebody_hits_8h)$gene_name
#   ) %>%
#   dplyr::select(
#     gene_id,
#     gene_name,
#     seqnames,
#     start,
#     end,
#     logCPM,
#     logFC,
#     hmPValue,
#     FDR = hmPValue_fdr,
#     FDR_sig
#   )

# readr::write_tsv(df_genebody_8h_out, "xchip_8h_gene_body.tsv")
```


FDR does not give any significant regions. Under a coordinated global shift, the signal may be spread across many features rather than concentrated in a subset of highly significant responders. Standard FDR correction controls false discoveries at the individual-feature level and can therefore have limited sensitivity for detecting broad, low-amplitude changes

So to overcome this, we generate a significance score metric, combing log2FC and pvalue to evalue the significance of differential signal of Pol II.
Significance Score=∣logFC∣×(−log10​(PValue)) - no directions
Signed Significance Score=logFC×(−log10​(PValue))

read in the 90mins TSS files
```{r}
diffsig_tss_90min <- readRDS("df_tss_90min.rds")
#diffsig_tss_90min$signed_significance_score <- sign(diffsig_tss_90min$logFC) * (-log10(diffsig_tss_90min$PValue))

#only annotate the significant ones based on new rank
diffsig_tss_90min <- diffsig_tss_90min %>%
  dplyr::mutate(
    signed_significance_score = ifelse(
      PValue < 0.05,
      sign(logFC) * (-log10(PValue)),
      NA
    )
  )

```

find overlaps with what genes are these
```{r}
library(GenomicRanges)
library(rtracklayer)
# Convert your TSS table to GRanges
tss_90_gr <- GRanges(
  seqnames = sub(":.*", "", diffsig_tss_90min$range),
  ranges = IRanges(
    start = as.numeric(sub(".*:", "", sub("-.*", "", diffsig_tss_90min$range))),
    end = as.numeric(sub(".*-", "", diffsig_tss_90min$range))
  )
)

# Find overlaps
hits_tss_90min <- findOverlaps(tss_90_gr, gene.gr )

# Add gene information
diffsig_tss_90min$gene_id <- NA
diffsig_tss_90min$gene_name <- NA

diffsig_tss_90min$gene_id[queryHits(hits_tss_90min)] <- gene.gr$gene_id[subjectHits(hits_tss_90min)]
diffsig_tss_90min$gene_name[queryHits(hits_tss_90min)] <- gene.gr$gene_name[subjectHits(hits_tss_90min)]



#export the output for the data
write_csv(diffsig_tss_90min, "diffsig_tss_90min_significant_score_metadata.csv")
```
regenerate the MA plot with the new significance score rank
the mean significance score was calculated after the plotting code then used to refine the annotation of the plot

```{r}
#extract the relevant mean significance scorw
tss_90min_gene_summary <- dfs_diff_sig_mean$tss_90min

#plot the refined MA plot
ggplot(diffsig_tss_90min, aes(x = logCPM, y = logFC)) +

  # Points coloured by signed significance score
  geom_point(
    aes(color = signed_significance_score),
    alpha = 0.7,
    size = 0.8
  ) +

  geom_hline(
    yintercept = 0,
    linetype = "dotted",
    color = "black"
  ) +

  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "grey80",
    name = "Signed\nsignificance score"
  ) +

  scale_y_continuous(
    limits = c(-2, 2)
  ) +

  theme(
    legend.position = "right",
    legend.direction = "vertical",
    plot.title = element_text(hjust = 0.5, size = 12),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    panel.background = element_rect(fill = "white", colour = NA)
  ) +

  labs(
    x = "logCPM",
    y = "log2 Fold Change",
    title = "Differential signal TSS MA plot with signed significance score (90 min vs 0 h)"
  ) +

 annotate(
  "text",
  x = Inf,
  y = Inf,
  hjust = 1.1,
  vjust = 1.5,
  label = paste0(
    "Significant windows: ",
    sum(diffsig_tss_90min$PValue < 0.05, na.rm = TRUE),
    " / ",
    nrow(diffsig_tss_90min),
    "\n",
    "Genes after merging windows: ",
    sum(!is.na(tss_90min_gene_summary$mean_signed_significance_score)),
    "\n",
    "Genes with blue score: ",
    sum(tss_90min_gene_summary$colour == "blue", na.rm = TRUE),
    "\n",
    "Genes with red score: ",
    sum(tss_90min_gene_summary$colour == "red", na.rm = TRUE)
  )
)+
ggdag::remove_grid()
```

read in the 8hmins TSS files
```{r}
diffsig_tss_8h <- readRDS("df_tss_8h.rds")
#diffsig_tss_8h$signed_significance_score <- sign(diffsig_tss_8h$logFC) * (-log10(diffsig_tss_8h$PValue))

#only annotate the significant ones based on new rank
diffsig_tss_8h <- diffsig_tss_8h %>%
  dplyr::mutate(
    signed_significance_score = ifelse(
      PValue < 0.05,
      sign(logFC) * (-log10(PValue)),
      NA
    )
  )

```

find overlaps with what genes are these
```{r}
library(GenomicRanges)
library(rtracklayer)
# Convert your TSS table to GRanges
tss_8h_gr <- GRanges(
  seqnames = sub(":.*", "", diffsig_tss_8h$range),
  ranges = IRanges(
    start = as.numeric(sub(".*:", "", sub("-.*", "", diffsig_tss_8h$range))),
    end = as.numeric(sub(".*-", "", diffsig_tss_8h$range))
  )
)

# Find overlaps
hits_tss_8h <- findOverlaps(tss_8h_gr, gene.gr )

# Add gene information
diffsig_tss_8h$gene_id <- NA
diffsig_tss_8h$gene_name <- NA

diffsig_tss_8h$gene_id[queryHits(hits_tss_8h)] <- gene.gr$gene_id[subjectHits(hits_tss_8h)]
diffsig_tss_8h$gene_name[queryHits(hits_tss_8h)] <- gene.gr$gene_name[subjectHits(hits_tss_8h)]

#export the output for the data
write_csv(diffsig_tss_8h, "diffsig_tss_8h_significant_score_metadata.csv")
```
regenerate the MA plot with the new significance score rank

```{r}

#extract the relevant mean significant score
tss_8h_gene_summary <- dfs_diff_sig_mean$tss_8h

#plot the refined MA plot
ggplot(diffsig_tss_8h, aes(x = logCPM, y = logFC)) +

  # Points coloured by signed significance score
  geom_point(
    aes(color = signed_significance_score),
    alpha = 0.7,
    size = 0.8
  ) +

  geom_hline(
    yintercept = 0,
    linetype = "dotted",
    color = "black"
  ) +

  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "grey80",
    name = "Signed\nsignificance score"
  ) +

  scale_y_continuous(
    limits = c(-2, 2)
  ) +

  theme(
    legend.position = "right",
    legend.direction = "vertical",
    plot.title = element_text(hjust = 0.5, size = 12),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    panel.background = element_rect(fill = "white", colour = NA)
  ) +

  labs(
    x = "logCPM",
    y = "log2 Fold Change",
    title = "Differential signal TSS MA plot with signed significance score (8h vs 0 h)"
  ) +

 annotate(
  "text",
  x = Inf,
  y = Inf,
  hjust = 1.1,
  vjust = 1.5,
  label = paste0(
    "Significant windows: ",
    sum(diffsig_tss_8h$PValue < 0.05, na.rm = TRUE),
    " / ",
    nrow(diffsig_tss_8h),
    "\n",
    "Genes after merging windows: ",
    sum(!is.na(tss_8h_gene_summary$mean_signed_significance_score)),
    "\n",
    "Genes with blue score: ",
    sum(tss_8h_gene_summary$colour == "blue", na.rm = TRUE),
    "\n",
    "Genes with red score: ",
    sum(tss_8h_gene_summary$colour == "red", na.rm = TRUE)
  )
)+
ggdag::remove_grid()
```


Repeat the same for gene body
read in the 90mins genebody files
```{r}
diffsig_genebody_90min <- readRDS("df_genebody_90min.rds")
#diffsig_genebody_90min$signed_significance_score <- sign(diffsig_genebody_90min$logFC) * (-log10(diffsig_genebody_90min$PValue))

#only annotate the significant ones based on new rank
diffsig_genebody_90min <- diffsig_genebody_90min %>%
  dplyr::mutate(
    signed_significance_score = ifelse(
      PValue < 0.05,
      sign(logFC) * (-log10(PValue)),
      NA
    )
  )

```

find overlaps with what genes are these
```{r}
library(GenomicRanges)
library(rtracklayer)



# Convert your genebody table to GRanges
genebody_90_gr <- GRanges(
  seqnames = sub(":.*", "", diffsig_genebody_90min$range),
  ranges = IRanges(
    start = as.numeric(sub(".*:", "", sub("-.*", "", diffsig_genebody_90min$range))),
    end = as.numeric(sub(".*-", "", diffsig_genebody_90min$range))
  )
)

# Find overlaps
hits_genebody_90min <- findOverlaps(genebody_90_gr, gene.gr )

# Add gene information
diffsig_genebody_90min$gene_id <- NA
diffsig_genebody_90min$gene_name <- NA

diffsig_genebody_90min$gene_id[queryHits(hits_genebody_90min)] <- gene.gr$gene_id[subjectHits(hits_genebody_90min)]
diffsig_genebody_90min$gene_name[queryHits(hits_genebody_90min)] <- gene.gr$gene_name[subjectHits(hits_genebody_90min)]

#export the output for the data
write_csv(diffsig_genebody_90min, "diffsig_genebody_90min_significant_score_metadata.csv")
```
regenerate the MA plot with the new significance score rank
the mean significance score was calculated after the MAplots

```{r}
#extract the relevant mean significance score
genebody_90min_gene_summary <- dfs_diff_sig_mean$genebody_90min

#plot the refined MA plot
ggplot(diffsig_genebody_90min, aes(x = logCPM, y = logFC)) +

  # Points coloured by signed significance score
  geom_point(
    aes(color = signed_significance_score),
    alpha = 0.7,
    size = 0.8
  ) +

  geom_hline(
    yintercept = 0,
    linetype = "dotted",
    color = "black"
  ) +

  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "grey80",
    name = "Signed\nsignificance score"
  ) +

  scale_y_continuous(
    limits = c(-2, 2)
  ) +

  theme(
    legend.position = "right",
    legend.direction = "vertical",
    plot.title = element_text(hjust = 0.5, size = 12),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    panel.background = element_rect(fill = "white", colour = NA)
  ) +

  labs(
    x = "logCPM",
    y = "log2 Fold Change",
    title = "Differential signal genebody MA plot with signed significance score (90 min vs 0 h)"
  ) +

  annotate(
  "text",
  x = Inf,
  y = Inf,
  hjust = 1.1,
  vjust = 1.5,
  label = paste0(
    "Significant windows: ",
    sum(diffsig_genebody_90min$PValue < 0.05, na.rm = TRUE),
    " / ",
    nrow(diffsig_genebody_90min),
    "\n",
    "Genes after merging windows: ",
    sum(!is.na(genebody_90min_gene_summary$mean_signed_significance_score)),
    "\n",
    "Genes with blue score: ",
    sum(genebody_90min_gene_summary$colour == "blue", na.rm = TRUE),
    "\n",
    "Genes with red score: ",
    sum(genebody_90min_gene_summary$colour == "red", na.rm = TRUE)
  )
) +

ggdag::remove_grid()
```

read in the 8hmins genebody files
```{r}
diffsig_genebody_8h <- readRDS("df_genebody_8h.rds")
#diffsig_genebody_8h$signed_significance_score <- sign(diffsig_genebody_8h$logFC) * (-log10(diffsig_genebody_8h$PValue))

#only annotate the significant ones based on new rank
diffsig_genebody_8h <- diffsig_genebody_8h %>%
  dplyr::mutate(
    signed_significance_score = ifelse(
      PValue < 0.05,
      sign(logFC) * (-log10(PValue)),
      NA
    )
  )

```

find overlaps with what genes are these
```{r}
library(GenomicRanges)
library(rtracklayer)
# Convert your genebody table to GRanges
genebody_8h_gr <- GRanges(
  seqnames = sub(":.*", "", diffsig_genebody_8h$range),
  ranges = IRanges(
    start = as.numeric(sub(".*:", "", sub("-.*", "", diffsig_genebody_8h$range))),
    end = as.numeric(sub(".*-", "", diffsig_genebody_8h$range))
  )
)

# Find overlaps
hits_genebody_8h <- findOverlaps(genebody_8h_gr, gene.gr )

# Add gene information
diffsig_genebody_8h$gene_id <- NA
diffsig_genebody_8h$gene_name <- NA

diffsig_genebody_8h$gene_id[queryHits(hits_genebody_8h)] <- gene.gr$gene_id[subjectHits(hits_genebody_8h)]
diffsig_genebody_8h$gene_name[queryHits(hits_genebody_8h)] <- gene.gr$gene_name[subjectHits(hits_genebody_8h)]

#export the output for the data
write_csv(diffsig_genebody_8h, "diffsig_genebody_8h_significant_score_metadata.csv")
```
regenerate the MA plot with the new significance score rank
for annotation of number of genes, use mean significance score rank (calculated after this block of code then added to the plotting)

```{r}

#extract the relevant mean significant score
genebody_8h_gene_summary <- dfs_diff_sig_mean$genebody_8h

#plot the refined MA plot
ggplot(diffsig_genebody_8h, aes(x = logCPM, y = logFC)) +

  # Points coloured by signed significance score
  geom_point(
    aes(color = signed_significance_score),
    alpha = 0.7,
    size = 0.8
  ) +

  geom_hline(
    yintercept = 0,
    linetype = "dotted",
    color = "black"
  ) +

  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "grey80",
    name = "Signed\nsignificance score"
  ) +

  scale_y_continuous(
    limits = c(-2, 2)
  ) +

  theme(
    legend.position = "right",
    legend.direction = "vertical",
    plot.title = element_text(hjust = 0.5, size = 12),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    panel.background = element_rect(fill = "white", colour = NA)
  ) +

  labs(
    x = "logCPM",
    y = "log2 Fold Change",
    title = "Differential signal genebody MA plot with signed significance score (8h vs 0 h)"
  ) +

annotate(
  "text",
  x = Inf,
  y = Inf,
  hjust = 1.1,
  vjust = 1.5,
  label = paste0(
    "Significant windows: ",
    sum(diffsig_genebody_8h$PValue < 0.05, na.rm = TRUE),
    " / ",
    nrow(diffsig_genebody_8h),
    "\n",
    "Genes after merging windows: ",
    sum(!is.na(genebody_8h_gene_summary$mean_signed_significance_score)),
    "\n",
    "Genes with blue score: ",
    sum(genebody_8h_gene_summary$colour == "blue", na.rm = TRUE),
    "\n",
    "Genes with red score: ",
    sum(genebody_8h_gene_summary$colour == "red", na.rm = TRUE)
  )
) +
ggdag::remove_grid()
```

---
Extended analysis from differential signal
---

Given that we have observed genes with higher pol II occupancy at 90min/8h compared to 0h and vice versa, we decided to investigate further on the properties of genes which showed differential signal.

generate the metadata this time to include both the 90min and the 8h dataset, along with information such as pausing index,baseMean and gene length
```{r}
metadata_alltimepoints <- cov_boxplot_dfs_norm %>% dplyr::select(gene_id,hgnc_symbol,dataset, condition,coverage.fold)  %>% 
  pivot_wider(
    names_from = condition,
    values_from = coverage.fold
  ) %>%
  left_join(pausingratio_csv_expanded_0h_baseline, #checked that the basemean are quite similar between 90min and 8h in pausingratio_csv_expanded, this is just the property of the pausing ratio at 0h (native)
            by = c("hgnc_symbol" = "gene.name")) %>%  mutate(length_category = case_when(
    length >= 1000  & length < 5000  ~ "1_5kb",
    length >= 5000  & length < 10000 ~ "5_10kb",
    length >= 10000 & length < 25000 ~ "10_25kb",
    length >= 25000 & length < 50000 ~ "25_50kb",
    length >= 50000 & length < 75000 ~ "50_75kb",
    length >= 75000 ~ "75kb_plus",
    TRUE ~ NA_character_
  ))
```

append the differential signal analysis data
```{r}

#list all the differential signal datasets together
dfs_diff_sig <- list(
  tss_90min = diffsig_tss_90min,
  tss_8h = diffsig_tss_8h,
  genebody_90min = diffsig_genebody_90min,
  genebody_8h = diffsig_genebody_8h
)

#get average signed_significance as the signed significance score is per sliding window so we get an average to merge
dfs_diff_sig_mean <- lapply(dfs_diff_sig, function(df) {
  df %>%
    group_by(gene_id) %>%
    summarise(
      mean_signed_significance_score = mean(signed_significance_score, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      mean_signed_significance_score = ifelse(
        is.nan(mean_signed_significance_score),
        NA_real_,
        mean_signed_significance_score
      ),
      colour = case_when(
        is.na(mean_signed_significance_score) ~ "grey",
        mean_signed_significance_score > 0 ~ "red",
        mean_signed_significance_score < 0 ~ "blue",
        TRUE ~ "grey"
      ),
      colour = replace_na(colour, "grey")
    )
})
#> lapply(dfs_diff_sig_mean, function(df) {
#+     df %>%
#+         filter(colour %in% c("red", "blue")) %>%
#+         count(colour)
#+ })
#$tss_90min
# A tibble: 2 × 2
#  colour     n
#  <chr>  <int>
#1 blue    1443
#2 red      153

#$tss_8h
# A tibble: 2 × 2
#  colour     n
#  <chr>  <int>
#1 blue    1154
#2 red      179

#$genebody_90min
# A tibble: 2 × 2
#  colour     n
#  <chr>  <int>
#1 blue    6865
#2 red     3741

#$genebody_8h
# A tibble: 2 × 2
#  colour     n
#  <chr>  <int>
#1 blue    8307
#2 red     2499


result <- metadata_alltimepoints

for (nm in names(dfs_diff_sig_mean)) {
  result <- result %>%
    dplyr::left_join(
      dfs_diff_sig_mean[[nm]] %>%
        dplyr::rename(
          !!paste0("mean_signed_significance_score_", nm) := mean_signed_significance_score,
          !!paste0("colour_", nm) := colour
        ),
      by = "gene_id"
    )
}
```

now plot the boxplots of 1) basemeanexpression, pausing index(log2), and gene length

```{r}
diffsig_plot_df <- result %>%
  pivot_longer(
    cols = starts_with("colour_"),
    names_to = "comparison",
    values_to = "colour"
  ) %>%
  mutate(
    comparison = str_remove(comparison, "^colour_"),
    timepoint = if_else(str_detect(comparison, "90min"), "90 min", "8 h"),
    comparison_dataset = case_when(
      str_detect(comparison, "^tss") ~ "TSS",
      str_detect(comparison, "^genebody") ~ "Gene Body"
    )
  ) %>%
  filter(dataset == comparison_dataset) %>% 
  dplyr::filter(!is.na(colour))


#make sure the plots are in the correct order
diffsig_plot_df <- diffsig_plot_df %>%
  filter(!is.na(colour)) %>%
  mutate(
    timepoint = factor(timepoint, levels = c("90 min", "8 h")),
    dataset = factor(dataset, levels = c("TSS", "Gene Body"))
  )
#basemean
diffsig_baseMean_plot <- ggplot(diffsig_plot_df,
       aes(x = colour,
           y = log2 (baseMean_90min_0h),
           fill = colour)) +
  geom_boxplot(outlier.alpha = 0.3) +
  theme_bw() +facet_grid(dataset ~ timepoint)+
  labs(
    x = "Differential significance",
    y = "log2 baseMean expression"
  ) +
  scale_fill_identity()

#log2 pausing ratio
diffsig_log2_pausingratio_plot <- ggplot(diffsig_plot_df,
       aes(x = colour,
           y = log2_pausing_ratio,
           fill = colour)) +
  geom_boxplot(outlier.alpha = 0.3) +
  theme_bw() +facet_grid(dataset ~ timepoint) +
  labs(
    x = "Differential significance",
    y = "log2 Pausing ratio"
  ) +
  scale_fill_identity()

#genelength
diffsig_log2_genelength_plot <- ggplot(diffsig_plot_df,
       aes(x = colour,
           y = log2(length),
           fill = colour)) +
  geom_boxplot(outlier.alpha = 0.3) +
  theme_bw() +facet_grid(dataset ~ timepoint)+
  labs(
    x = "Differential significance",
    y = "log2 gene length"
  ) +
  scale_fill_identity()

 
 #combine the plots together by patchwork

diffsig_baseMean_plot + diffsig_log2_pausingratio_plot  + diffsig_log2_genelength_plot
```

---
Scatter plots 
---

plot scatter plot of log2FC of TSS vs gene body for both timepoints : 90min and 8h, compared to 0h

```{r}

#add the mean significance score and colour information back to the individual dataframe
dfs_diff_sig_joined <- purrr::map2(dfs_diff_sig, dfs_diff_sig_mean, ~
  .x %>%
    left_join(
      .y %>%
        select(gene_id, mean_signed_significance_score, colour),
      by = "gene_id"
    )
)

#combine all thedataframes together
dfs_diff_sig_joined_combined <- bind_rows(dfs_diff_sig_joined, .id = "source") 

#generate the data required to plot TSS vs genebody scatterplot
scatter_diffsig_plot_df <- dfs_diff_sig_joined_combined %>%
  mutate(
    region = case_when(
      grepl("^tss", source) ~ "TSS",
      grepl("^genebody", source) ~ "Genebody"
    ),
    time = case_when(
      grepl("90min", source) ~ "90min",
      grepl("8h", source) ~ "8h"
    )
  ) %>%
  group_by(gene_id, gene_name, time, region) %>%
  summarise(
  logFC = mean(logFC, na.rm = TRUE),
  colour = colour[1],
  .groups = "drop"
) %>%
  pivot_wider(
    names_from = region,
    values_from = c(logFC, colour)
  ) %>%
  mutate(
    colour_combo = paste(colour_TSS, colour_Genebody, sep = "_")
  ) %>% mutate(
    direction_class = case_when(
      colour_combo %in% c("red_red", "blue_blue") ~ "Same direction",
      colour_combo %in% c("red_blue", "blue_red") ~ "Opposite direction",
      colour_combo %in% c("red_grey", "blue_grey") ~ "TSS only",
      colour_combo %in% c("grey_red", "grey_blue") ~ "Gene body only",
      colour_combo == "grey_grey" ~ "No change",
      TRUE ~ "Other"
    )
  )

combo_cols <- c(
  "Same direction"    = "purple",
  "Opposite direction" = "orange",
  "TSS only"          = "red",
  "Gene body only"    = "blue",
  "No change"         = "grey70"
)


#reorder the levels
scatter_diffsig_plot_df <- scatter_diffsig_plot_df %>%
  mutate(
    time = factor(time, levels = c("90min", "8h"))
  )


#plot the scatter plot of comparisons
TSSvsgenebody_scatter_plot <- ggplot(scatter_diffsig_plot_df,
       aes(x = logFC_TSS,
           y = logFC_Genebody,
           colour = direction_class)) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey70") +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey70") +
  geom_point(alpha = 0.8, size = 2) +
  facet_wrap(~time) +
  scale_colour_manual(values = combo_cols, na.value = "black") +
  labs(
    x = "TSS log2FC",
    y = "Gene body log2FC",
    colour = "Significance"
  ) +
  theme_bw() + ggdag::remove_grid()

#generate the more refined statistics of TSS:Genebody
colour_combo_table <- scatter_diffsig_plot_df %>%
  group_by(colour_combo, time) %>%
  summarise(n = n(), .groups = "drop")

View(colour_combo_table)

#density plots for the TSS vs genebody scatter plots
TSSvsgenebody_scatter_densityplot <- ggplot(
  scatter_diffsig_plot_df,
  aes(x = logFC_TSS, y = logFC_Genebody)
) + scale_x_continuous(breaks = c(-2,-1,0,1,2)) + 
  scale_y_continuous(breaks = c(-2,-1,0,1,2))+
  geom_bin2d(bins = 80) +
  scale_fill_viridis_c(name = "Gene count") +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey70") +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey70") +
  facet_wrap(~time) +
  labs(
    x = "TSS log2FC",
    y = "Gene body log2FC"
  ) +
  theme_bw()+ggdag::remove_grid()


#with correlation coeefficient
library(ggplot2)
library(ggpubr)
library(ggdag)

TSSvsgenebody_scatter_densityplot_corr <- ggplot(
  scatter_diffsig_plot_df,
  aes(x = logFC_TSS, y = logFC_Genebody)
) +
  scale_x_continuous(
    breaks = c(-2, -1, 0, 1, 2)
  ) +
  scale_y_continuous(
    breaks = c(-2, -1, 0, 1, 2)
  ) +
  geom_bin2d(bins = 80) +
  scale_fill_viridis_c(name = "Gene count") +
  geom_smooth(
    method = "lm",
    colour = "red",
    linewidth = 1,
    se = FALSE
  ) +
  stat_cor(
    method = "pearson",
    aes(label = paste(..r.label.., ..p.label.., sep = "~`,`~")),
    label.x.npc = "left",
    label.y.npc = "top",
    size = 5
  ) +
  geom_hline(
    yintercept = 0,
    linetype = 2,
    colour = "grey70"
  ) +
  geom_vline(
    xintercept = 0,
    linetype = 2,
    colour = "grey70"
  ) +
  facet_wrap(~time) +
  labs(
    x = "TSS log2 Fold Change",
    y = "Gene body log2 Fold Change"
  ) +
  theme_bw() +
  ggdag::remove_grid() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text = element_text(size = 12, colour = "black"),
    axis.title = element_text(size = 14),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11)
  )

```

scatter plot of log2FC to gene length , but highlighting the differential signal status. note now the log2FC is refers to the average foldchange of TSS and genebody respectively. 

```{r}
#add pausing ratio and gene length to the TSS/genebody log2FC from diffsig
scatter_diffsig_plot_df2 <- scatter_diffsig_plot_df %>% left_join(
    hg38.genes %>% dplyr::select(gene_id, width),
    by = "gene_id"
  ) %>% dplyr::rename("length" ="width") %>%
  left_join(pausingratio_csv_expanded_0h_baseline %>% dplyr::select(gene.name, log2_pausing_ratio), 
            by = c("gene_name" = "gene.name")) %>%
  mutate(
    time = factor(time, levels = c("90min","8h"))
  )



#plot the scatter plots
# TSS
TSS_logFCvsgenelength_plot <- scatter_diffsig_plot_df2 %>%
  filter(!is.na(length), length > 0) %>%
  mutate(
    length_bin = cut(
      length,
      breaks = quantile(length, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE),
      include.lowest = TRUE,
      labels = c("Short", "Medium", "Long")
    )
  ) %>%
  ggplot(aes(x = length, y = logFC_TSS, colour = colour_Genebody)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0) +
  scale_x_log10() +
  scale_colour_manual(
    values = c("blue" = "blue", "red" = "red", "grey" = "grey"),
    na.value = "grey"
  ) +
  facet_grid(time ~ length_bin, scales = "free_x") +
  theme_bw() +
  ggdag::remove_grid() +
  labs(x = "log10 (gene length)")

#density map overlap for logFC_TSS vs length 
TSS_logFCvsgenelength_densityplot <- scatter_diffsig_plot_df2 %>%
  filter(!is.na(length), !is.na(logFC_TSS), length > 0) %>%
  mutate(
    length_bin = cut(
      length,
      breaks = quantile(length, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE),
      include.lowest = TRUE,
      labels = c("Short", "Medium", "Long")
    )
  ) %>%
  ggplot(aes(x = length, y = logFC_TSS)) +
  geom_bin2d(bins = 80) +
  geom_hline(yintercept = 0) +
  scale_x_log10() +
  scale_fill_viridis_c() +
  facet_grid(time ~ length_bin, scales = "free_x") +
  labs(
    x = "log10 (gene length)",
    y = "logFC_TSS",
    fill = "Gene density"
  ) +
  theme_bw() +
  ggdag::remove_grid()

#combine the two genebody plots together using patchwork
TSS_logFCvsgenelength_plot/TSS_logFCvsgenelength_densityplot

# Gene body
genebody_logFCvsgenelength_plot <- scatter_diffsig_plot_df2 %>%
  filter(!is.na(length), length > 0) %>%
  mutate(
    length_bin = cut(
      length,
      breaks = quantile(length, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE),
      include.lowest = TRUE,
      labels = c("Short", "Medium", "Long")
    )
  ) %>%
  ggplot(aes(x = length, y = logFC_Genebody, colour = colour_Genebody)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0) +
  scale_x_log10() +
  scale_colour_manual(
    values = c("blue" = "blue", "red" = "red", "grey" = "grey"),
    na.value = "grey"
  ) +
  facet_grid(time ~ length_bin, scales = "free_x")+
  theme_bw() +
  ggdag::remove_grid() + labs( x = "log10 (gene length)")


#density map overlap for logFC_genebody vs length 
genebody_logFCvsgenelength_densityplot <- scatter_diffsig_plot_df2 %>%
  filter(!is.na(length), !is.na(logFC_Genebody), length > 0) %>%
  mutate(
    length_bin = cut(
      length,
      breaks = quantile(length, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE),
      include.lowest = TRUE,
      labels = c("Short", "Medium", "Long")
    )
  ) %>%
  ggplot(aes(x = length, y = logFC_Genebody)) +
  geom_bin2d(bins = 80) +
  geom_hline(yintercept = 0) +
  scale_x_log10() +
  scale_fill_viridis_c() +
  facet_grid(time ~ length_bin, scales = "free_x") +
  labs(
    x = "log10 (gene length)",
    y = "logFC_Genebody",
    fill = "Gene density"
  ) +
  theme_bw() +
  ggdag::remove_grid()

#combine the two genebody plots together using patchwork
genebody_logFCvsgenelength_plot/genebody_logFCvsgenelength_densityplot
```

 Now plot log2FC of TSS/genebody over pausing index
 
```{r}
#plot the scatter plots
# TSS
TSS_logFCvspausingindex_plot <- scatter_diffsig_plot_df2  %>%
  ggplot(aes(x = log2_pausing_ratio, y = logFC_TSS, colour = colour_Genebody)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0) +
  scale_colour_manual(
    values = c("blue" = "blue", "red" = "red", "grey" = "grey"),
    na.value = "grey"
  ) +
  facet_grid(time ~ .)+
  theme_bw() +
  ggdag::remove_grid() + labs( x = "log2 (pausing index at 0h)")

#density map overlap for logFC_TSS vs length 
TSS_logFCvspausingindex_densityplot <- scatter_diffsig_plot_df2 %>%
  ggplot(aes(x = log2_pausing_ratio, y = logFC_TSS)) +
  geom_bin2d(bins = 80) +
  geom_hline(yintercept = 0) +
  scale_fill_viridis_c() +
  labs(
    x = "log2 (pausing index at 0h)",
    y = "logFC_TSS",
    fill = "Gene density"
  ) +facet_grid(time ~ .)+
  theme_bw() +
  ggdag::remove_grid()

#combine the two genebody plots together using patchwork
TSS_logFCvspausingindex_plot/TSS_logFCvspausingindex_densityplot

# Gene body
genebody_logFCvspausingindex_plot <- scatter_diffsig_plot_df2 %>%
  ggplot(aes(x = log2_pausing_ratio, y = logFC_Genebody, colour = colour_Genebody)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0) +
  scale_colour_manual(
    values = c("blue" = "blue", "red" = "red", "grey" = "grey"),
    na.value = "grey"
  ) +facet_grid(time ~ .)+
  theme_bw() +
  ggdag::remove_grid() + labs( x = "log2 (pausing index at 0h)")


#density map overlap for logFC_genebody vs length 
genebody_logFCvspausingindex_densityplot <- scatter_diffsig_plot_df2 %>%
  ggplot(aes(x = log2_pausing_ratio, y = logFC_Genebody)) +
  geom_bin2d(bins = 80) +
  geom_hline(yintercept = 0) +
  scale_fill_viridis_c() +
  labs(
    x = "log2 (pausing index at 0h)",
    y = "logFC_Genebody",
    fill = "Gene density"
  ) +facet_grid(time ~ .)+
  theme_bw() +
  ggdag::remove_grid()

#combine the two genebody plots together using patchwork
genebody_logFCvspausingindex_plot/genebody_logFCvspausingindex_densityplot
```
 
 
---
promoter analysis of diff sig data
---
 
add the promoter type information onto the differential signal analysis data. As the gene sets enriched for TSS and gene body are different so we will investigate the promoter enrichment for both.
```{r}

#keeping only the TSS
#diffsig_plot_df_TSS_only <- diffsig_plot_df %>% filter(dataset == "TSS")


#calculate the percentage of promoter types from the differential signal TSS dataset
diffsig_plot_df_perc <- diffsig_plot_df %>%
  mutate(
    promoter_type = coalesce(
      promoter_type_NA,
      promoter_type_1,
      promoter_type_2,
      promoter_type_3,
      promoter_type_4
    )
  ) %>%
  count(timepoint, dataset, colour, promoter_type) %>%
  group_by(timepoint, colour) %>%
  mutate(percent = round(n / sum(n) * 100, 2)) %>%
  ungroup() %>%
  dplyr::select(timepoint, dataset, promoter_type, colour, percent) %>%
  pivot_wider(
    names_from = colour,
    values_from = percent,
    values_fill = 0
  )

#now calculate the overall promoter percentage
allgenes_promoter_perc <- all_promoters_enum %>%
  count(promoter_type) %>%
  mutate(
    percent = sprintf("%.2f", n / sum(n) * 100)
  ) %>%
  arrange(desc(n))

#combine the summary tables
promoter_summary <-diffsig_plot_df_perc %>%
  left_join(allgenes_promoter_perc, by = "promoter_type") %>% 
  dplyr::select(-n) %>%dplyr::rename("all genes" = "percent") %>% 
  dplyr::select(promoter_type, timepoint,dataset,last_col(), everything())

#calculate promoter enrichment heatmap
diff_sig_promoter_enrichment_heatmap_df <- promoter_summary %>% 
  filter(promoter_type != "no promoter information") %>%
  mutate(across(c(`all genes`, blue, grey, red), as.numeric)) %>%
  pivot_longer(
    cols = c(blue, grey, red),
    names_to = "colour",
    values_to = "percent"
  ) %>%
  group_by(timepoint, dataset, promoter_type) %>%
  mutate(
    enrichment = percent / `all genes`,
    log2_enrichment = log2(enrichment)
  ) %>%
  ungroup() %>%
  mutate(
    promoter_type = factor(
      promoter_type,
      levels = c(
        "GC_box",
        "CCAAT",
        "Initiator",
        "TATA"
      )
    )
  )



#plot promoter enrichment heatmap (all )
promoter_enrichment_heatmap <-ggplot(
  diff_sig_promoter_enrichment_heatmap_df,
  aes(colour, promoter_type, fill = log2_enrichment)
) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(log2_enrichment, 2)), size = 4) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    name = "log2\nEnrichment"
  ) +
  facet_grid(dataset ~ timepoint) +
  coord_equal() +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.title = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )



#calculate using the grey genes as the background instead of all genes
diff_sig_promoter_enrichment_heatmap_df_greybg  <- promoter_summary %>% 
  filter(promoter_type != "no promoter information") %>%
  mutate(across(c(grey, blue, red), as.numeric)) %>%
  mutate(
    blue_enrichment = blue / grey,
    red_enrichment  = red / grey,
    grey_enrichment = 1
  ) %>%
  pivot_longer(
    cols = ends_with("_enrichment"),
    names_to = "colour",
    values_to = "enrichment"
  ) %>%
  mutate(
    colour = sub("_enrichment", "", colour),
    log2_enrichment = log2(enrichment)
  )


promoter_enrichment_greybg_heatmap <- ggplot(
  diff_sig_promoter_enrichment_heatmap_df_greybg %>%
    filter(colour != "grey"),
  aes(colour, promoter_type, fill = log2_enrichment)
) +
  geom_tile(color = "white") +
  geom_text(
    aes(label = round(log2_enrichment, 2)),
    size = 4
  ) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    name = "log2\nEnrichment\n(vs grey)"
  ) +
  facet_grid(dataset ~ timepoint) +
  coord_equal() +
  theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.title = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```
 
