---
title: "Chip_seq data analysis"
output: html_document
date: "2025-06-20"
---

This project is to analyse chipseq data for the Degron RPB1 project in collaboration with Scott and Alex from UNSW.

---
Differential signal analysis using extraChips
---

Given that Pol II is a combination of broad and narrow peaks,  unlike histones and TFs which is either broad or narrow, I have decided to go with differential signal analysis which goes by sliding window approach


Tutorial for extraCHIPS : https://bioconductor.org/packages//release/bioc/vignettes/extraChIPs/inst/doc/differential_signal_sliding.html
This part requires more than >50GB of RAM, so this part was deployed on the cluster via R

```{r, run_csaw.R}
#!/usr/bin/env Rscript

library(csaw)
library(extraChIPs)
library(Rsamtools)
library(rtracklayer)
library(glue)
library(here)
library(stringr)
library(magrittr)
library(tidyverse)
library(GenomeInfoDb)

# BAM files
bam_files <- c(
  "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/D8L4Y_bam/D8L4Y_0h_rep1.mLb.clN.sorted.bam",
  "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/D8L4Y_bam/D8L4Y_0h_rep2.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/D8L4Y_bam/D8L4Y_0h_rep3.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/D8L4Y_bam/D8L4Y_90min_rep1.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/D8L4Y_bam/D8L4Y_90min_rep2.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/D8L4Y_bam/D8L4Y_90min_rep3.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/D8L4Y_bam/D8L4Y_8h_rep1.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/D8L4Y_bam/D8L4Y_8h_rep2.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/INP_0h_rep1.mLb.clN.sorted.bam",
  "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/INP_0h_rep2.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/INP_0h_rep3.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/INP_90min_rep1.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/INP_90min_rep2.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/INP_90min_rep3.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/INP_8h_rep1.mLb.clN.sorted.bam",
 "/lmb/home/png/cephfs2/Degron_Scott/Output/chipseq_out/bamfiles/INP_8h_rep2.mLb.clN.sorted.bam"

)

bfl <- BamFileList(bam_files)
names(bfl) <- tools::file_path_sans_ext(basename(bam_files))

# Seqinfo (keeps only the key chromosomes, as the data wats generated from nf-core so no chr prefix)
sq <- seqinfo(BamFile(bam_files[1]))
sq <- keepSeqlevels(sq, c(as.character(1:22), "X", "Y"))

# Grey/black lists
blacklist <- rtracklayer::import.bed(
  "/lmb/home/png/cephfs2/Genomes/merged_hg38_sacCer3/hg38-blacklist.v2.diffsig.bed"
)
blacklist <- keepSeqlevels(blacklist, seqlevels(sq), pruning.mode="coarse")

# Read parameters
rp <- readParam(
  pe = "none",
  dedup = TRUE,
  discard = blacklist
)

# Count windows
wincounts <- windowCounts(
  bam.files = bfl,
  spacing = 40,
  width = 120,
  ext = 200,
  filter = length(bfl),
  param = rp
)


# Save output
#saveRDS(wincounts, file = "Degron_chip_wincounts.rds")
sample <- data.frame(samples = c("D8L4Y_0h_rep1.mLb.clN.sorted",
                                 "D8L4Y_0h_rep2.mLb.clN.sorted",
                                 "D8L4Y_0h_rep3.mLb.clN.sorted",
                                 "D8L4Y_90min_rep1.mLb.clN.sorted",
                                 "D8L4Y_90min_rep2.mLb.clN.sorted",
                                 "D8L4Y_90min_rep3.mLb.clN.sorted",
                                 "D8L4Y_8h_rep1.mLb.clN.sorted",
                                 "D8L4Y_8h_rep2.mLb.clN.sorted",
                                 "INP_0h_rep1.mLb.clN.sorted",
                                 "INP_0h_rep2.mLb.clN.sorted",
                                 "INP_0h_rep3.mLb.clN.sorted",
                                 "INP_90min_rep1.mLb.clN.sorted",
                                 "INP_90min_rep2.mLb.clN.sorted",
                                 "INP_90min_rep3.mLb.clN.sorted",
                                 "INP_8h_rep1.mLb.clN.sorted",
                                 "INP_8h_rep2.mLb.clN.sorted"),
                     group = c("D8L4Y_0h",
                                 "D8L4Y_0h",
                                 "D8L4Y_0h",
                                 "D8L4Y_90min",
                                 "D8L4Y_90min",
                                 "D8L4Y_90min",
                                 "D8L4Y_8h",
                                 "D8L4Y_8h",
                                 "INP_0h",
                                 "INP_0h",
                                 "INP_0h",
                                 "INP_90min",
                                 "INP_90min",
                                 "INP_90min",
                                 "INP_8h",
                                 "INP_8h"))
#add information to the data

colData(wincounts) <- colData(wincounts) %>%
  as_tibble(rownames = "samples") %>% 
  left_join(sample, by = "samples") %>% 
  dplyr::select(
    samples, all_of(colnames(colData(wincounts))), group
  ) %>%
  DataFrame(row.names = .$samples)


  #import consensus narrow peaks file
#peaks <- import("/cephfs2/png/Degron_Scott/Output/chipseq_out/D8L4Y.consensus_peaks.bed")

#replace peaks with genes ie regions of interest
regions <-  import("/cephfs2/png/Genomes/hg38/Homo_sapiens.GRCh38.113_genes_.gtf")
#start the filtering process, match it to the sample's respective input files
samples_list <- colnames(wincounts)

#remove sample3 input for 8h as it is not used for the analysis
samples_list <- samples_list[!samples_list %in% c(
  "INP_8h_rep3.mLb.clN.sorted"
)]

samples_list <- samples_list[!grepl("^INP_", samples_list)]

filtered_list <- list()

for (s in samples_list) {

  idx <- match(s, colnames(wincounts))
  if (is.na(idx)) next

  bg_sample <- sub("^D8L4Y", "INP", s)
  bg_idx <- match(bg_sample, colnames(wincounts))
  if (is.na(bg_idx)) next

  filtered_list[[s]] <- dualFilter(
    x   = wincounts[, idx],
    bg  = wincounts[, bg_idx],
    ref = regions,
    q   = 0.6
  )
  #combined_counts <- do.call(cbind,
  #lapply(filtered_list, assay)
#)
}

saveRDS(filtered_list, "Degron_chip_filtered_counts_regions.RDS")

cat("Finished successfully\n")
```


read in the filtered counts RDS object and proceed with differential signal analysis using sliding window

```{r, diff_signal analysis}
#!/usr/bin/env Rscript

library(csaw)
library(extraChIPs)
library(Rsamtools)
library(rtracklayer)
library(glue)
library(here)
library(stringr)
library(magrittr)
library(tidyverse)
library(GenomeInfoDb)
library(ggplot2)
library(viridis)
library(MASS)
library(ggdag)

filtered_wincounts <- readRDS("Degron_chip_filtered_counts_regions.RDS")

# find common genomic windows
common_ids <- Reduce(intersect,
                     lapply(filtered_wincounts, function(x) {
                         rr <- rowRanges(x)
                         paste0(seqnames(rr), ":", start(rr), "-", end(rr))
                     }))

# subset all objects to same rows
filtered_common <- lapply(filtered_wincounts, function(se) {

    rr <- rowRanges(se)
    ids <- paste0(seqnames(rr), ":", start(rr), "-", end(rr))

    se[ids %in% common_ids, ]
})

# reorder rows identically
filtered_common <- lapply(filtered_common, function(se) {

    rr <- rowRanges(se)
    ids <- paste0(seqnames(rr), ":", start(rr), "-", end(rr))

    se[match(common_ids, ids), ]
})

# combine samples
combined <- do.call(cbind, filtered_common)

#plot assay densities with the common ones
# Assaydensity <- plotAssayDensities(combined,
#                    assay = "logCPM",
#                    colour = "sample") +
#   theme_bw()

#PCA
# PCA <- plotAssayPCA(combined, "logCPM", colour = "group", label = "sample") +
#   theme_bw()
 
#  ggsave(
#     filename = "PCA_extraChips.pdf",
#     plot = PCA,
#     width = 6,
#     height = 4
# )
#RLE
# RLE_plot <-  plotAssayRle(combined, assay = "logCPM", fill = "sample") +
#   geom_hline(yintercept = 0, linetype = 2, colour = "grey") +
#   ggtitle("RLE: Across All Samples") +
#   theme_bw() +
#    theme(axis.text.x = element_text(angle = 90, hjust = 1))

 
#  ggsave(
#     filename = "RLE_extraChips.pdf",
#     plot = RLE_plot,
#     width = 6,
#     height = 4
# )


# Differential region analysis and significance-aware merging
# ____________________________________________________________
#
# This workflow processes differential binding results separately
# for each experimental contrast (90 min vs 0 h and 8 h vs 0 h).
#
# 1. Significance definition:
#    For each fitted model, regions are classified as significant
#    based on an FDR threshold (< 0.05).
#
# 2. Conditional region merging:
#    - If no significant regions are detected, the original
#      window-level results are retained and placeholder values
#      are assigned for merged p-values.
#    - If significant regions exist, adjacent significant windows
#      are merged using a harmonic mean p-value (HMP) approach
#      within a specified genomic distance (merge_within = 120 bp).
#
# 3. Post-processing:
#    - Merged regions are annotated with differential status
#      (via addDiffStatus).
#    - This allows downstream visualization and interpretation
#      of region-level differential signal.
#
# 4. Purpose:
#    This step reduces redundancy from overlapping sliding windows
#    and enables identification of broader genomic regions of
#    consistent differential signal across conditions.
#
# NOTE:
# Each contrast (fit_90, fit_8h) is processed independently to
# preserve condition-specific biological signals.
#
# ============================================================
library(scales)
X <- model.matrix(~group, data = colData(combined))

#X
#                                (Intercept) groupD8L4Y_8h groupD8L4Y_90min
#D8L4Y_0h_rep1.mLb.clN.sorted              1             0                0
#D8L4Y_0h_rep2.mLb.clN.sorted              1             0                0
#D8L4Y_0h_rep3.mLb.clN.sorted              1             0                0
#D8L4Y_90min_rep1.mLb.clN.sorted           1             0                1
#D8L4Y_90min_rep2.mLb.clN.sorted           1             0                1
#D8L4Y_90min_rep3.mLb.clN.sorted           1             0                1
#D8L4Y_8h_rep1.mLb.clN.sorted              1             1                0
#D8L4Y_8h_rep2.mLb.clN.sorted              1             1                0
#attr(,"assign")
#[1] 0 1 1
#attr(,"contrasts")
#attr(,"contrasts")$group
#[1] "contr.treatment"




#differential signal of 90min vs 0h 
fit_90 <- fitAssayDiff(
  combined,
  design = X,
  coef = "groupD8L4Y_90min",
  fc = 1.0, #no foldchange implied, unless i use lfc
  asRanges = TRUE
)

#differential signal of 8h vs 0h
fit_8h <- fitAssayDiff(
  combined,
  design = X,
  coef = "groupD8L4Y_8h",
  fc = 1.0, #no foldchange implied, unless i use lfc
  asRanges = TRUE
)


# all_plot <- fit_gr %>% 
#   as_tibble() %>% 
#   ggplot(aes(logCPM, logFC)) +
#   geom_point(alpha = 0.6) +
#   geom_smooth(se = FALSE, method = "loess") +
#   geom_label(
#     aes(label = label),
#     data = . %>% 
#       summarise(
#         n = dplyr::n(),
#         logCPM = max(logCPM) - 0.15 * diff(range(logCPM)),
#         logFC = max(logFC) - 0.05 * diff(range(logFC)),
#         label = glue("{comma(n)}\nSliding Windows")
#       )
#   ) +
#   ylim(range(fit_gr$logFC)) +
#   ggtitle("MA Plot: All Retained Sliding Windows")



#plot the differential binding result, but did not work because checked there is no signal
fit_90$sig <- fit_90$FDR < 0.05
fit_8h$sig <- fit_8h$FDR < 0.05


if (!any(fit_90$sig, na.rm = TRUE)) {

    results_90_xchip <- fit_90
    mcols(results_90_xchip)$hmPValue <- NA
    mcols(results_90_xchip)$hmPValue_fdr <- NA

} else {

    results_90_xchip <- mergeByHMP(
        fit_90,
        sig_col = "sig",
        merge_within = 120
    ) %>%
        addDiffStatus(
            sig_col = "sig",
            drop = TRUE
        )
}

if (!any(fit_8h$sig, na.rm = TRUE)) {

    results_8h_xchip <- fit_8h
    mcols(results_8h_xchip)$hmPValue <- NA
    mcols(results_8h_xchip)$hmPValue_fdr <- NA

} else {

    results_8h_xchip <- mergeByHMP(
        fit_8h,
        sig_col = "sig",
        merge_within = 120
    ) %>%
        addDiffStatus(
            sig_col = "sig",
            drop = TRUE
        )
}

#check for number of merged_windows which are less than 0.05

table(results_8h_xchip$FDR < 0.05)
table(results_90_xchip$FDR < 0.05)

#cross check the stats before merging using mergeHPM
#> table(results_8h_xchip$FDR < 0.05)
#  FALSE 
# 4013881 
#table(results_90_xchip$FDR < 0.05)
#  FALSE 
# 4013881 

#90min vs 0h
df_90 <- results_90_xchip %>%
  as_tibble() %>%
  mutate(FDR_sig = hmPValue_fdr < 0.05)


#generate the all regions MA plots
FDR_plot_90 <- ggplot(df_90, aes(logCPM, logFC)) +

  geom_point(alpha = 0.5, colour = "grey70") +

  geom_point(
    data = df_90 %>% dplyr::filter(FDR_sig),
    colour = "red",
    alpha = 0.8
  ) +

   geom_hline(yintercept = 0) +
  
  scale_y_continuous(limits = c(-2,2), breaks = c(-2, -1, 0, 1, 2)) +
  annotate(
    "text",
    x = Inf, y = Inf,
    hjust = 1.1, vjust = 1.5,
    label = paste0(
      "FDR < 0.05: ",
      sum(df_90$FDR_sig, na.rm = TRUE),
      " regions\n(out of ",
      nrow(df_90),
      ")"
    )
  ) +
  theme(
    legend.position = "right",
    legend.direction = "vertical",
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    strip.text.x = element_text(size = 12),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 12),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
   panel.background = element_rect(fill = "white", colour = NA)
  ) +
  ggtitle("MA Plot: differential signal analysis 90 min vs 0h") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  remove_grid()

#8h vs 0h
df_8h <- results_8h_xchip %>%
  as_tibble() %>%
  mutate(FDR_sig = hmPValue_fdr < 0.05)


#generate the all regions MA plots
FDR_plot_8h <- ggplot(df_8h, aes(logCPM, logFC)) +

  geom_point(alpha = 0.5, colour = "grey70") +

  geom_point(
    data = df_8h %>% dplyr::filter(FDR_sig),
    colour = "red",
    alpha = 0.8
  ) +

  geom_hline(yintercept = 0) +
scale_y_continuous(limits = c(-2,2),breaks = c(-2, -1, 0, 1, 2)) +
  annotate(
    "text",
    x = Inf, y = Inf,
    hjust = 1.1, vjust = 1.5,
    label = paste0(
      "FDR < 0.05: ",
      sum(df_8h$FDR_sig, na.rm = TRUE),
      " regions\n(out of ",
      nrow(df_8h),
      ")"
    )
  ) +
  theme(
    legend.position = "right",
    legend.direction = "vertical",
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    strip.text.x = element_text(size = 12),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 12),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
   panel.background = element_rect(fill = "white", colour = NA)
  ) +
  ggtitle("MA Plot: differential signal analysis 8h vs 0h") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  remove_grid()

ggsave("MA_xchips_90min_vs_0h.png", FDR_plot_90, width = 8, height = 6, dpi = 300)
ggsave("MA_xchips_8h_vs_0h.png", FDR_plot_8h, width = 8, height = 6, dpi = 300)


#now do by TSS
gtf.dat <- import("/cephfs2/png/Genomes/hg38/Homo_sapiens.GRCh38.113.gtf")
gene.gr <- gtf.dat[gtf.dat$type %in% "gene",]
names(gene.gr) <- gene.gr$gene_id
gene.gr$gene.width <- GenomicRanges::width(gene.gr)
gene.gr$gene_name_column <- gene.gr$gene_name

gene.gr <- gene.gr[width(gene.gr) >1000]
gene.gr <- gene.gr[gene.gr$gene_biotype %in% "protein_coding"]
gene.gr <- gene.gr[countOverlaps(gene.gr,gene.gr) == 1]


#make sure gene.gr is > 1kb and only has non-overlapping protein coding genes
tss <- resize(gene.gr, width = 1, fix = "start")
tss_500 <- promoters(tss, upstream = 30, downstream = 300)
#flank(tss, width = 500, start = TRUE)
tss_hits <- subsetByOverlaps(results_90_xchip, tss_500)

df_tss <- as_tibble(tss_hits) %>%
  mutate(FDR_sig = FDR < 0.05)


n_sig2 <- sum(df_tss$FDR_sig, na.rm = TRUE)
n_tot2 <- nrow(df_tss)


# #plot the TSS only MA density plot
MAplot_TSS_90min <- ggplot(df_tss, aes(x = logCPM, y = logFC)) +

  # optional faint points (same as your res_df2 plot)
  geom_point(alpha = 0.3, size = 0.8, color = "grey") +

  # 2D density (matches your style)
  stat_density_2d(
    aes(fill = after_stat(level), alpha = after_stat(level)),
    geom = "polygon",
    contour = TRUE,
    bins = 20
  ) +

  # FDR significant points
  geom_point(
    data = df_tss %>% dplyr::filter(FDR_sig),
    color = "red",
    size = 0.8
  ) +

  geom_hline(yintercept = 0, linetype = "dotted", color = "black") +

  scale_fill_viridis_c(option = "magma", name = "Density") +
  scale_alpha(range = c(0.2, 0.8), guide = "none") +

  scale_y_continuous(limits = c(-1.5, 1.5),breaks=c(-1.5,-1,0,1,1.5)) +

    theme(
    legend.position = "right",
    legend.direction = "vertical",
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    strip.text.x = element_text(size = 12),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 12),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    panel.background = element_rect(fill = "white", colour = NA)
  )  +
  labs(
    x = "logCPM",
    y = "log2 Fold Change",
    title = "differential signal TSS MA plot with density (90min vs 0h)"
  ) +
  annotate(
  "text",
  x = Inf, y = Inf,
  hjust = 1.1, vjust = 1.5,
  label = paste0(
    "FDR < 0.05: ",
    sum(df_tss$FDR_sig, na.rm = TRUE),
    " / ",
    nrow(df_tss)
  )
) +
  remove_grid()

#8h TSS±500bp
tss_hits_8h <- subsetByOverlaps(results_8h_xchip, tss_500)

df_tss_8h <- as_tibble(tss_hits_8h) %>%
  mutate(FDR_sig = FDR < 0.05)

n_sig <- sum(df_tss_8h$FDR_sig, na.rm = TRUE)
n_tot <- nrow(df_tss_8h)


#generate the TSS desnity MA plot
MAplot_TSS_8h <- ggplot(df_tss_8h, aes(x = logCPM, y = logFC)) +

  # optional faint points (same as your res_df2 plot)
  geom_point(alpha = 0.3, size = 0.8, color = "grey") +

  # 2D density (matches your style)
  stat_density_2d(
    aes(fill = after_stat(level), alpha = after_stat(level)),
    geom = "polygon",
    contour = TRUE,
    bins = 20
  ) +

  # FDR significant points
  geom_point(
    data = df_tss_8h %>% dplyr::filter(FDR_sig),
    color = "red",
    size = 0.8
  ) +

  geom_hline(yintercept = 0, linetype = "dotted", color = "black") +

  scale_fill_viridis_c(option = "magma", name = "Density") +
  scale_alpha(range = c(0.2, 0.8), guide = "none") +

  scale_y_continuous(limits = c(-1.5, 1.5),breaks=c(-1.5,-1,0,1,1.5)) +

    theme(
    legend.position = "right",
    legend.direction = "vertical",
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    strip.text.x = element_text(size = 12),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 12),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    panel.background = element_rect(fill = "white", colour = NA)
  )  +
  labs(
    x = "logCPM",
    y = "log2 Fold Change",
    title = "differential signal TSS MA plot with density (8h vs 0h)"
  ) +
  annotate(
  "text",
  x = Inf, y = Inf,
  hjust = 1.1, vjust = 1.5,
  label = paste0(
    "FDR < 0.05: ",
    sum(df_tss_8h$FDR_sig, na.rm = TRUE),
    " / ",
    nrow(df_tss_8h)
  )
) +
  remove_grid()

ggsave("MA_xchips_TSS_90min_vs_0h.png", MAplot_TSS_90min, width = 8, height = 6, dpi = 300)
ggsave("MA_xchips_TSS_8h_vs_0h.png", MAplot_TSS_8h, width = 8, height = 6, dpi = 300)


###repeat the same for genebody
hg38.gene.body.bed <-  read.table(
    "/cephfs2/png/Genomes/hg38/hg38.gene.body.bed",
    sep = "\t",
    header = TRUE,
    stringsAsFactors = FALSE
)

genebody <- GenomicRanges::makeGRangesFromDataFrame(
  hg38.gene.body.bed ,
  seqnames.field = "chr",
  start.field    = "start",
  end.field      = "end",
  keep.extra.columns = TRUE
)

seqlevels(genebody) <- sub("^chr", "", seqlevels(genebody))
seqnames(genebody) <- sub("^chr", "", seqnames(genebody))


#90mins vs 0h
genebody_hits <- subsetByOverlaps(results_90_xchip, genebody)

df_genebody <- as_tibble(genebody_hits) %>%
  mutate(FDR_sig = FDR < 0.05)


#check the significance statistics
n_sig3 <- sum(df_genebody$FDR_sig, na.rm = TRUE)
n_tot3 <- nrow(df_genebody)


#generate the density MA plot
MAplot_genebody_90mins <- ggplot(df_genebody, aes(x = logCPM, y = logFC)) +

  # optional faint points (same as your res_df2 plot)
  geom_point(alpha = 0.3, size = 0.8, color = "grey") +

  # 2D density (matches your style)
  stat_density_2d(
    aes(fill = after_stat(level), alpha = after_stat(level)),
    geom = "polygon",
    contour = TRUE,
    bins = 20
  ) +

  # FDR significant points
  geom_point(
    data = df_genebody %>% dplyr::filter(FDR_sig),
    color = "red",
    size = 0.8
  ) +

  geom_hline(yintercept = 0, linetype = "dotted", color = "black") +

  scale_fill_viridis_c(option = "magma", name = "Density") +
  scale_alpha(range = c(0.2, 0.8), guide = "none") +

  scale_y_continuous(limits = c(-2, 2)) +

   theme(
    legend.position = "right",
    legend.direction = "vertical",
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    strip.text.x = element_text(size = 12),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 12),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    panel.background = element_rect(fill = "white", colour = NA)
  )  +
  labs(
    x = "logCPM",
    y = "log2 Fold Change",
    title = "differential signal gene body MA plot with density (90min vs 0h)"
  ) +
  annotate(
  "text",
  x = Inf, y = Inf,
  hjust = 1.1, vjust = 1.5,
  label = paste0(
    "FDR < 0.05: ",
    sum(df_genebody$FDR_sig, na.rm = TRUE),
    " / ",
    nrow(df_genebody)
  )
) +
  remove_grid()


#8h vs 0h
genebody_hits_8h <- subsetByOverlaps(results_8h_xchip, genebody)

df_genebody_8h <- as_tibble(genebody_hits_8h) %>%
  mutate(FDR_sig = FDR < 0.05)

#check the significance statistics
n_sig4 <- sum(df_genebody_8h$FDR_sig, na.rm = TRUE)
n_tot4 <- nrow(df_genebody_8h)


#generate the density MA plot
MAplot_genebody_8h <- ggplot(df_genebody_8h, aes(x = logCPM, y = logFC)) +

  # optional faint points (same as your res_df2 plot)
  geom_point(alpha = 0.3, size = 0.8, color = "grey") +

  # 2D density (matches your style)
  stat_density_2d(
    aes(fill = after_stat(level), alpha = after_stat(level)),
    geom = "polygon",
    contour = TRUE,
    bins = 20
  ) +

  # FDR significant points
  geom_point(
    data = df_genebody_8h %>% dplyr::filter(FDR_sig),
    color = "red",
    size = 0.8
  ) +

  geom_hline(yintercept = 0, linetype = "dotted", color = "black") +

  scale_fill_viridis_c(option = "magma", name = "Density") +
  scale_alpha(range = c(0.2, 0.8), guide = "none") +

  scale_y_continuous(limits = c(-2, 2)) +

    theme(
    legend.position = "right",
    legend.direction = "vertical",
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    strip.text.x = element_text(size = 12),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 12),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    panel.background = element_rect(fill = "white", colour = NA)
  )  +
  labs(
    x = "logCPM",
    y = "log2 Fold Change",
    title = "differential signal gene body MA plot with density (8h vs 0h)"
  ) +
  annotate(
  "text",
  x = Inf, y = Inf,
  hjust = 1.1, vjust = 1.5,
  label = paste0(
    "FDR < 0.05: ",
    sum(df_genebody_8h$FDR_sig, na.rm = TRUE),
    " / ",
    nrow(df_genebody_8h)
  )
) +
  remove_grid()

  
ggsave("MA_xchips_genebody_90min_vs_0h.png", MAplot_genebody_90mins, width = 8, height = 6, dpi = 300)
ggsave("MA_xchips_genebody_8h_vs_0h.png", MAplot_genebody_8h, width = 8, height = 6, dpi = 300)

saveRDS(df_tss, "df_tss_90min.rds")
saveRDS(df_tss_8h, "df_tss_8h.rds")

saveRDS(df_genebody, "df_genebody_90min.rds")
saveRDS(df_genebody_8h, "df_genebody_8h.rds")

#move the stats genartion part to local Rstudio
#generating tsv reports
# df_90_out <- results_90_xchip %>%
#   as_tibble() %>%
#   dplyr::select(seqnames, start, end, logCPM, logFC, hmPValue, hmPValue_fdr)

# readr::write_tsv(df_90_out, "xchip_90min_vs_0h_all_regions.tsv")

# df_8h_out <- results_8h_xchip %>%
#   as_tibble() %>%
#   dplyr::select(seqnames, start, end, logCPM, logFC, hmPValue, hmPValue_fdr)

# readr::write_tsv(df_8h_out, "xchip_8h_vs_0h_all_regions.tsv")

# df_tss_out <- df_tss %>%
#   mutate(
#     gene_id = mcols(tss_hits)$gene_id,
#     gene_name = mcols(tss_hits)$gene_name
#   ) %>%
#   dplyr::select(
#     gene_id,
#     gene_name,
#     seqnames,
#     start,
#     end,
#     logCPM,
#     logFC,
#     hmPValue,
#     FDR = hmPValue_fdr,
#     FDR_sig
#   )

# readr::write_tsv(df_tss_out, "xchip_90min_TSS.tsv")

# df_tss_8h_out <- df_tss_8h %>%
#   mutate(
#     gene_id = mcols(tss_hits_8h)$gene_id,
#     gene_name = mcols(tss_hits_8h)$gene_name
#   ) %>%
#   dplyr::select(
#     gene_id,
#     gene_name,
#     seqnames,
#     start,
#     end,
#     logCPM,
#     logFC,
#     hmPValue,
#     FDR = hmPValue_fdr,
#     FDR_sig
#   )

# readr::write_tsv(df_tss_8h_out, "xchip_8h_TSS.tsv")


# df_genebody_90min_out <- df_genebody %>%
#   mutate(
#     gene_id = mcols(genebody_hits)$gene_id,
#     gene_name = mcols(genebody_hits)$gene_name
#   ) %>%
#   dplyr::select(
#     gene_id,
#     gene_name,
#     seqnames,
#     start,
#     end,
#     logCPM,
#     logFC,
#     hmPValue,
#     FDR = hmPValue_fdr,
#     FDR_sig
#   )

# readr::write_tsv(df_genebody_90min_out, "xchip_90min_gene_body.tsv")


# df_genebody_8h_out <- df_genebody_8h %>%
#   mutate(
#     gene_id = mcols(genebody_hits_8h)$gene_id,
#     gene_name = mcols(genebody_hits_8h)$gene_name
#   ) %>%
#   dplyr::select(
#     gene_id,
#     gene_name,
#     seqnames,
#     start,
#     end,
#     logCPM,
#     logFC,
#     hmPValue,
#     FDR = hmPValue_fdr,
#     FDR_sig
#   )

# readr::write_tsv(df_genebody_8h_out, "xchip_8h_gene_body.tsv")
```


FDR does not give any significant regions. Under a coordinated global shift, the signal may be spread across many features rather than concentrated in a subset of highly significant responders. Standard FDR correction controls false discoveries at the individual-feature level and can therefore have limited sensitivity for detecting broad, low-amplitude changes

So to overcome this, we generate a significance score metric, combing log2FC and pvalue to evalue the significance of differential signal of Pol II.
Significance Score=∣logFC∣×(−log10​(PValue)) - no directions
Signed Significance Score=logFC×(−log10​(PValue))

read in the 90mins TSS files
```{r}
diffsig_tss_90min <- readRDS("df_tss_90min.rds")
#diffsig_tss_90min$signed_significance_score <- sign(diffsig_tss_90min$logFC) * (-log10(diffsig_tss_90min$PValue))

#only annotate the significant ones based on new rank
diffsig_tss_90min <- diffsig_tss_90min %>%
  dplyr::mutate(
    signed_significance_score = ifelse(
      PValue < 0.05,
      sign(logFC) * (-log10(PValue)),
      NA
    )
  )

```

find overlaps with what genes are these
```{r}
library(GenomicRanges)
library(rtracklayer)
# Convert your TSS table to GRanges
tss_90_gr <- GRanges(
  seqnames = sub(":.*", "", diffsig_tss_90min$range),
  ranges = IRanges(
    start = as.numeric(sub(".*:", "", sub("-.*", "", diffsig_tss_90min$range))),
    end = as.numeric(sub(".*-", "", diffsig_tss_90min$range))
  )
)

# Find overlaps
hits_tss_90min <- findOverlaps(tss_90_gr, gene.gr )

# Add gene information
diffsig_tss_90min$gene_id <- NA
diffsig_tss_90min$gene_name <- NA

diffsig_tss_90min$gene_id[queryHits(hits_tss_90min)] <- gene.gr$gene_id[subjectHits(hits_tss_90min)]
diffsig_tss_90min$gene_name[queryHits(hits_tss_90min)] <- gene.gr$gene_name[subjectHits(hits_tss_90min)]

#export the output for the data
write_csv(diffsig_tss_90min, "diffsig_tss_90min_significant_score_metadata.csv")
```
regenerate the MA plot with the new significance score rank

```{r}
ggplot(diffsig_tss_90min, aes(x = logCPM, y = logFC)) +

  # Points coloured by signed significance score
  geom_point(
    aes(color = signed_significance_score),
    alpha = 0.7,
    size = 0.8
  ) +

  geom_hline(
    yintercept = 0,
    linetype = "dotted",
    color = "black"
  ) +

  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "grey80",
    name = "Signed\nsignificance score"
  ) +

  scale_y_continuous(
    limits = c(-2, 2)
  ) +

  theme(
    legend.position = "right",
    legend.direction = "vertical",
    plot.title = element_text(hjust = 0.5, size = 12),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    panel.background = element_rect(fill = "white", colour = NA)
  ) +

  labs(
    x = "logCPM",
    y = "log2 Fold Change",
    title = "Differential signal TSS MA plot with signed significance score (90 min vs 0 h)"
  ) +

  annotate(
    "text",
    x = Inf,
    y = Inf,
    hjust = 1.1,
    vjust = 1.5,
    label = paste0(
      "P-value < 0.05 windows: ",
      sum(diffsig_tss_90min$PValue < 0.05, na.rm = TRUE),
      " / ",
      nrow(diffsig_tss_90min),
      "\n",
      "P-value < 0.05 genes: ",
      length(unique(
        diffsig_tss_90min$gene_name[
          diffsig_tss_90min$PValue < 0.05
        ]
      )),
      " / ",
      length(unique(
        diffsig_tss_90min$gene_name
      ))
    )
  ) +

  remove_grid()
```

read in the 8hmins TSS files
```{r}
diffsig_tss_8h <- readRDS("df_tss_8h.rds")
#diffsig_tss_8h$signed_significance_score <- sign(diffsig_tss_8h$logFC) * (-log10(diffsig_tss_8h$PValue))

#only annotate the significant ones based on new rank
diffsig_tss_8h <- diffsig_tss_8h %>%
  dplyr::mutate(
    signed_significance_score = ifelse(
      PValue < 0.05,
      sign(logFC) * (-log10(PValue)),
      NA
    )
  )

```

find overlaps with what genes are these
```{r}
library(GenomicRanges)
library(rtracklayer)
# Convert your TSS table to GRanges
tss_8h_gr <- GRanges(
  seqnames = sub(":.*", "", diffsig_tss_8h$range),
  ranges = IRanges(
    start = as.numeric(sub(".*:", "", sub("-.*", "", diffsig_tss_8h$range))),
    end = as.numeric(sub(".*-", "", diffsig_tss_8h$range))
  )
)

# Find overlaps
hits_tss_8h <- findOverlaps(tss_8h_gr, gene.gr )

# Add gene information
diffsig_tss_8h$gene_id <- NA
diffsig_tss_8h$gene_name <- NA

diffsig_tss_8h$gene_id[queryHits(hits_tss_8h)] <- gene.gr$gene_id[subjectHits(hits_tss_8h)]
diffsig_tss_8h$gene_name[queryHits(hits_tss_8h)] <- gene.gr$gene_name[subjectHits(hits_tss_8h)]

#export the output for the data
write_csv(diffsig_tss_8h, "diffsig_tss_8h_significant_score_metadata.csv")
```
regenerate the MA plot with the new significance score rank

```{r}
ggplot(diffsig_tss_8h, aes(x = logCPM, y = logFC)) +

  # Points coloured by signed significance score
  geom_point(
    aes(color = signed_significance_score),
    alpha = 0.7,
    size = 0.8
  ) +

  geom_hline(
    yintercept = 0,
    linetype = "dotted",
    color = "black"
  ) +

  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "grey80",
    name = "Signed\nsignificance score"
  ) +

  scale_y_continuous(
    limits = c(-2, 2)
  ) +

  theme(
    legend.position = "right",
    legend.direction = "vertical",
    plot.title = element_text(hjust = 0.5, size = 12),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    panel.background = element_rect(fill = "white", colour = NA)
  ) +

  labs(
    x = "logCPM",
    y = "log2 Fold Change",
    title = "Differential signal TSS MA plot with signed significance score (8h vs 0 h)"
  ) +

  annotate(
    "text",
    x = Inf,
    y = Inf,
    hjust = 1.1,
    vjust = 1.5,
    label = paste0(
      "P-value < 0.05 windows: ",
      sum(diffsig_tss_8h$PValue < 0.05, na.rm = TRUE),
      " / ",
      nrow(diffsig_tss_8h),
      "\n",
      "P-value < 0.05 genes: ",
      length(unique(
        diffsig_tss_8h$gene_name[
          diffsig_tss_8h$PValue < 0.05
        ]
      )),
      " / ",
      length(unique(
        diffsig_tss_8h$gene_name
      ))
    )
  ) +

  remove_grid()
```


Repeat the same for gene body
read in the 90mins genebody files
```{r}
diffsig_genebody_90min <- readRDS("df_genebody_90min.rds")
#diffsig_genebody_90min$signed_significance_score <- sign(diffsig_genebody_90min$logFC) * (-log10(diffsig_genebody_90min$PValue))

#only annotate the significant ones based on new rank
diffsig_genebody_90min <- diffsig_genebody_90min %>%
  dplyr::mutate(
    signed_significance_score = ifelse(
      PValue < 0.05,
      sign(logFC) * (-log10(PValue)),
      NA
    )
  )

```

find overlaps with what genes are these
```{r}
library(GenomicRanges)
library(rtracklayer)
# Convert your genebody table to GRanges
genebody_90_gr <- GRanges(
  seqnames = sub(":.*", "", diffsig_genebody_90min$range),
  ranges = IRanges(
    start = as.numeric(sub(".*:", "", sub("-.*", "", diffsig_genebody_90min$range))),
    end = as.numeric(sub(".*-", "", diffsig_genebody_90min$range))
  )
)

# Find overlaps
hits_genebody_90min <- findOverlaps(genebody_90_gr, gene.gr )

# Add gene information
diffsig_genebody_90min$gene_id <- NA
diffsig_genebody_90min$gene_name <- NA

diffsig_genebody_90min$gene_id[queryHits(hits_genebody_90min)] <- gene.gr$gene_id[subjectHits(hits_genebody_90min)]
diffsig_genebody_90min$gene_name[queryHits(hits_genebody_90min)] <- gene.gr$gene_name[subjectHits(hits_genebody_90min)]

#export the output for the data
write_csv(diffsig_genebody_90min, "diffsig_genebody_90min_significant_score_metadata.csv")
```
regenerate the MA plot with the new significance score rank

```{r}
ggplot(diffsig_genebody_90min, aes(x = logCPM, y = logFC)) +

  # Points coloured by signed significance score
  geom_point(
    aes(color = signed_significance_score),
    alpha = 0.7,
    size = 0.8
  ) +

  geom_hline(
    yintercept = 0,
    linetype = "dotted",
    color = "black"
  ) +

  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "grey80",
    name = "Signed\nsignificance score"
  ) +

  scale_y_continuous(
    limits = c(-2, 2)
  ) +

  theme(
    legend.position = "right",
    legend.direction = "vertical",
    plot.title = element_text(hjust = 0.5, size = 12),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    panel.background = element_rect(fill = "white", colour = NA)
  ) +

  labs(
    x = "logCPM",
    y = "log2 Fold Change",
    title = "Differential signal genebody MA plot with signed significance score (90 min vs 0 h)"
  ) +

  annotate(
    "text",
    x = Inf,
    y = Inf,
    hjust = 1.1,
    vjust = 1.5,
    label = paste0(
      "P-value < 0.05 windows: ",
      sum(diffsig_genebody_90min$PValue < 0.05, na.rm = TRUE),
      " / ",
      nrow(diffsig_genebody_90min),
      "\n",
      "P-value < 0.05 genes: ",
      length(unique(
        diffsig_genebody_90min$gene_name[
          diffsig_genebody_90min$PValue < 0.05
        ]
      )),
      " / ",
      length(unique(
        diffsig_genebody_90min$gene_name
      ))
    )
  ) +

  remove_grid()
```

read in the 8hmins genebody files
```{r}
diffsig_genebody_8h <- readRDS("df_genebody_8h.rds")
#diffsig_genebody_8h$signed_significance_score <- sign(diffsig_genebody_8h$logFC) * (-log10(diffsig_genebody_8h$PValue))

#only annotate the significant ones based on new rank
diffsig_genebody_8h <- diffsig_genebody_8h %>%
  dplyr::mutate(
    signed_significance_score = ifelse(
      PValue < 0.05,
      sign(logFC) * (-log10(PValue)),
      NA
    )
  )

```

find overlaps with what genes are these
```{r}
library(GenomicRanges)
library(rtracklayer)
# Convert your genebody table to GRanges
genebody_8h_gr <- GRanges(
  seqnames = sub(":.*", "", diffsig_genebody_8h$range),
  ranges = IRanges(
    start = as.numeric(sub(".*:", "", sub("-.*", "", diffsig_genebody_8h$range))),
    end = as.numeric(sub(".*-", "", diffsig_genebody_8h$range))
  )
)

# Find overlaps
hits_genebody_8h <- findOverlaps(genebody_8h_gr, gene.gr )

# Add gene information
diffsig_genebody_8h$gene_id <- NA
diffsig_genebody_8h$gene_name <- NA

diffsig_genebody_8h$gene_id[queryHits(hits_genebody_8h)] <- gene.gr$gene_id[subjectHits(hits_genebody_8h)]
diffsig_genebody_8h$gene_name[queryHits(hits_genebody_8h)] <- gene.gr$gene_name[subjectHits(hits_genebody_8h)]

#export the output for the data
write_csv(diffsig_genebody_8h, "diffsig_genebody_8h_significant_score_metadata.csv")
```
regenerate the MA plot with the new significance score rank

```{r}
ggplot(diffsig_genebody_8h, aes(x = logCPM, y = logFC)) +

  # Points coloured by signed significance score
  geom_point(
    aes(color = signed_significance_score),
    alpha = 0.7,
    size = 0.8
  ) +

  geom_hline(
    yintercept = 0,
    linetype = "dotted",
    color = "black"
  ) +

  scale_color_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "grey80",
    name = "Signed\nsignificance score"
  ) +

  scale_y_continuous(
    limits = c(-2, 2)
  ) +

  theme(
    legend.position = "right",
    legend.direction = "vertical",
    plot.title = element_text(hjust = 0.5, size = 12),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 12),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    panel.background = element_rect(fill = "white", colour = NA)
  ) +

  labs(
    x = "logCPM",
    y = "log2 Fold Change",
    title = "Differential signal genebody MA plot with signed significance score (8h vs 0 h)"
  ) +

  annotate(
    "text",
    x = Inf,
    y = Inf,
    hjust = 1.1,
    vjust = 1.5,
    label = paste0(
      "P-value < 0.05 windows: ",
      sum(diffsig_genebody_8h$PValue < 0.05, na.rm = TRUE),
      " / ",
      nrow(diffsig_genebody_8h),
      "\n",
      "P-value < 0.05 genes: ",
      length(unique(
        diffsig_genebody_8h$gene_name[
          diffsig_genebody_8h$PValue < 0.05
        ]
      )),
      " / ",
      length(unique(
        diffsig_genebody_8h$gene_name
      ))
    )
  ) +

  remove_grid()
```


