---
title: "snRNA and lncRNA metagene profiles for chipseq"
author: "Pei Qin (Sabrina) Ng"
date: "2026-07-30"
output: html_document
---

This is to analyse metagene plot for chipseq metagene as in chrome-extension://efaidnbmnnnibpcajpcglclefindmkaj/https://www.biorxiv.org/content/10.1101/2025.09.18.677100v1.full.pdf
---
Metagene plot analysis - lncRNA
---
Gent the lncRNA annotation
```{r}
#annotation file (the same as in ELOA P1)
#hg38.gtf <- read.table("/Users/png/Downloads/Homo_sapiens.GRCh38.113.gtf", header=FALSE, sep= "\t")

#generate non-overlapping gene annotation (using Granges, rtracklayer , and genomicfeatures)
gtf.dat <- import("/Users/png/Downloads/Homo_sapiens.GRCh38.113.gtf")
lncRNA.gr <- gtf.dat[gtf.dat$type  %in% "gene",]
lncRNA.gr$gene.width <- GenomicRanges::width(lncRNA.gr)
lncRNA.gr <- lncRNA.gr[lncRNA.gr$gene_biotype %in% "lncRNA"]

lncRNA.gr <- keepSeqlevels(lncRNA.gr,c(as.character(1:22),"X","Y"),pruning.mode="coarse")
#lncRNA.gr<- lncRNA.gr[width(lncRNA.gr) >1000]
#convert to dataframe from this step 9Granges can be problematic)
hg38.lncRNA <- as.data.frame(lncRNA.gr) %>% dplyr::rename("chr" = "seqnames") %>% mutate(chr = paste0("chr", chr))



# Filter lncRNA by strand
hg38.lncRNA.pos <- hg38.lncRNA %>% filter(strand=="+") 
hg38.lncRNA.neg <- hg38.lncRNA %>% filter(strand=="-")
 

# Define upstream and downstream flanking regions 
# Note that mutate() updates columns from left to right: it first moves ‘start’ 5 kb upstream, then sets ‘end’ 5 kb downstream of this new start. The resulting window (for upstream flanking regions) therefore runs from −5 kb to the TSS.
hg38.lncRNA.pos.upstream.flank <- hg38.lncRNA.pos %>% mutate(start=start-5000,end=start+5000) %>% filter(start>0)
hg38.lncRNA.neg.upstream.flank <- hg38.lncRNA.neg  %>% mutate(start=end,end=start+5000) %>% filter(start>0)
hg38.lncRNA.pos.downstream.flank <- hg38.lncRNA.pos %>% mutate(start=end,end=start+5000)
hg38.lncRNA.neg.downstream.flank <- hg38.lncRNA.neg  %>% mutate(start=start-5000,end=start+5000)
 
# Save files
write.table(hg38.lncRNA.pos, "hg38.lncRNA.pos.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.lncRNA.neg, "hg38.lncRNA.neg.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.lncRNA.pos.upstream.flank, "hg38.lncRNA.pos.upstream.flank.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.lncRNA.neg.upstream.flank, "hg38.lncRNA.neg.upstream.flank.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.lncRNA.pos.downstream.flank, "hg38.lncRNA.pos.downstream.flank.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.lncRNA.neg.downstream.flank, "hg38.lncRNA.neg.downstream.flank.bed", sep = "\t", quote = F, col.names = F, row.names = F)


```


After this use bedtools to split windows.

```{bash, chipseq_bedtools_makewindows.sh}

#tile the bed files 
bedtools makewindows -b hg38.lncRNA.pos.bed -n 60 -i winnum > hg38.pos.windows.60w.bed
bedtools makewindows -b hg38.lncRNA.neg.bed -n 60 -i winnum -reverse > hg38.neg.windows.60w.bed


#for upstream flanked
bedtools makewindows -b hg38.lncRNA.pos.upstream.flank.bed -n 20 -i winnum > hg38.pos.upstream.windows.20w.bed
bedtools makewindows -b hg38.lncRNA.neg.upstream.flank.bed -n 20 -i winnum -reverse > hg38.neg.upstream.windows.20w.bed

#for downstream flanked

bedtools makewindows -b hg38.lncRNA.pos.downstream.flank.bed -n 20 -i winnum > hg38.pos.downstream.windows.20w.bed
bedtools makewindows -b hg38.lncRNA.neg.downstream.flank.bed -n 20 -i winnum -reverse > hg38.neg.downstream.windows.20w.bed
# Combine the files as in the Star protocol (instead of R use awk so I can sort)

# Step 1: Combine gene tiles and add 20 to bin numbers (V4)
awk 'BEGIN{OFS="\t"} { $4 = $4 + 20; print }' hg38.pos.windows.60w.bed > lncRNA.tmp
awk 'BEGIN{OFS="\t"} { $4 = $4 + 20; print }' hg38.neg.windows.60w.bed >> lncRNA.tmp

# Step 2: Combine upstream flanks without change
cat hg38.pos.upstream.windows.20w.bed hg38.neg.upstream.windows.20w.bed > upstream.tmp


# Step 3: Combine downstream flanks and add 80 to bin numbers (V4)
awk 'BEGIN{OFS="\t"} { $4 = $4 + 80; print }' hg38.pos.downstream.windows.20w.bed > downstream.tmp
awk 'BEGIN{OFS="\t"} { $4 = $4 + 80; print }' hg38.neg.downstream.windows.20w.bed >> downstream.tmp

# Step 4: Combine all tiles
cat lncRNA.tmp upstream.tmp downstream.tmp > hg38.lncRNA.and.flanks.tiled.bed

# Step 5: Cleanup temporary files
rm lncRNA.tmp upstream.tmp downstream.tmp

# Sort
sort -k1,1 -k2,2n hg38.lncRNA.and.flanks.tiled.bed > hg38.lncRNA.and.flanks.tiled.sorted.bed

```

map it onto the binned, non-overlapping bedfiles

note: For this part of the script, use the same name bash script and run it on the cluster

```{bash, chipseq_array_bigwig.sh}

############################## generate big wig files using job array ###################

#Pre-script execution:
#Pre-script execution:
#Before running this script, run this code to generate file list with job arrays:
#ls ${FILESDIR}| rev | cut -c 12- | rev | uniq > chip_bw_list.txt


#these individual big wig file has been normalised using RPKM
#run UCSC bigwigMerge to combine the bigwigs (do all except for the rep3 8h - exclude from the study as we have verieid that it is a problematic sample, moved it to merged library)
bigWigMerge ${FILESDIR}/${PREFIX}_rep1.bigWig ${FILESDIR}/${PREFIX}_rep2.bigWig ${FILESDIR}/${PREFIX}_rep3.bigWig ${OUTDIR}/${PREFIX}_merged.bedGraph 

# Check for errors
if [ $? -ne 0 ]; then
    echo "bigwig merging failed for ${PREFIX}" >&2
    exit 1
else
    echo "bigwig merging succeeded for ${PREFIX}"
fi


#sort the bedgraphs
bedSort ${OUTDIR}/${PREFIX}_merged.bedGraph  ${OUTDIR}/${PREFIX}_merged_sorted.bedGraph

#convert the bedgraph to bigwig again
bedGraphToBigWig ${OUTDIR}/${PREFIX}_merged_sorted.bedGraph ${REFDIR}/homo_sapiens__GRCh38__release102.dna.fa.sizes ${OUTDIR}/${PREFIX}_merged.bw

# Check for errors
if [ $? -ne 0 ]; then
    echo "conversion to bigwig failed for ${PREFIX}" >&2
    exit 3
else
    echo "conversion to bigwig succeeded for ${PREFIX}"
fi

#filter for main chromosomes but also append "chr" to the chromosomes that were kept for mapping compatibility downstream
awk '$1 ~ /^(1?[0-9]|2[0-2]|X|Y|M)$/ { $1 = "chr"$1; print }' OFS='\t' "${OUTDIR}/${PREFIX}_merged_sorted.bedGraph" > "${OUTDIR}/${PREFIX}_merged_sorted.clean.bedGraph"
# Check for errors
if [ $? -ne 0 ]; then
    echo "bed cleaning new failed for ${PREFIX}" >&2
    exit 6
else
    echo "bed cleaning new succeeded for ${PREFIX}"
fi



###################################### DO NOT RUN this as part of the pipeline #####################
#blacklisted regions has been filtered out using samtools and bamtools according to the nextflowpipeline


#parse the defined bed files from gtfs and map it to the cleaned bed files (sorted using bedSort)
#bedtools map -a ./Datasets/POLII_ARMC5_8_24/Analysis/Windows/hg38.TSS.windows.100w.sorted.bed -b ./Datasets/POLII_ARMC5_8_24/bed_RPKM/K8_INP_merged.50nt.RPKM.blacklist.bed -c 4 -o mean -null 0 > ./Datasets/POLII_ARMC5_8_24/Analysis/Coverage/hg38.genes.K8_INP.fwd.coverage.bed
#bedtools map -a ./Datasets/POLII_ARMC5_8_24/Analysis/Windows/hg38.TSS.windows.100w.sorted.bed -b ./Datasets/POLII_ARMC5_8_24/bed_RPKM/K_INP_merged.50nt.RPKM.blacklist.bed -c 4 -o mean -null 0 > ./Datasets/POLII_ARMC5_8_24/Analysis/Coverage/hg38.genes.K_INP.fwd.coverage.bed

bedtools map -a ${OUTDIR}/hg38.genes.and.flanks.tiled.sorted.bed -b ${OUTDIR}/${PREFIX}_merged_sorted.clean.bedGraph -c 4 -o mean -null 0 > ${OUTDIR}/${PREFIX}_merged.clean.cov_redone.bed
# Check for errors
if [ $? -ne 0 ]; then
    echo "mapping to tiled gtf failed for ${PREFIX}" >&2
    exit 9
else
    echo "mapping to tiled gtf succeeded for ${PREFIX}"
fi

```


---
sRNA
---

```{r}
gtf.dat <- import("/Users/png/Downloads/Homo_sapiens.GRCh38.113.gtf")
srna.gr <- gtf.dat[gtf.dat$type  %in% "gene",]
srna.gr$gene.width <- GenomicRanges::width(srna.gr)
srna.gr<- srna.gr[srna.gr$gene_biotype %in% "snRNA"]
srna.gr <- keepSeqlevels(srna.gr,c(as.character(1:22),"X","Y"),pruning.mode="coarse")
srna.gr <- srna.gr[width(srna.gr ) >60]
#convert to dataframe from this step 9Granges can be problematic)
hg38.srna <- as.data.frame(srna.gr) %>% dplyr::rename("chr" = "seqnames") %>% mutate(chr = paste0("chr", chr))



# Filter srna by strand
hg38.srna.pos <- hg38.srna %>% filter(strand=="+") 
hg38.srna.neg <- hg38.srna %>% filter(strand=="-")
 

# Define upstream and downstream flanking regions 
# Note that mutate() updates columns from left to right: it first moves ‘start’ 5 kb upstream, then sets ‘end’ 5 kb downstream of this new start. The resulting window (for upstream flanking regions) therefore runs from −5 kb to the TSS.
hg38.srna.pos.upstream.flank <- hg38.srna.pos %>% mutate(start=start-5000,end=start+5000) %>% filter(start>0)
hg38.srna.neg.upstream.flank <- hg38.srna.neg  %>% mutate(start=end,end=start+5000) %>% filter(start>0)
hg38.srna.pos.downstream.flank <- hg38.srna.pos %>% mutate(start=end,end=start+5000)
hg38.srna.neg.downstream.flank <- hg38.srna.neg  %>% mutate(start=start-5000,end=start+5000)
 
# Save files
write.table(hg38.srna.pos, "hg38.srna.pos.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.srna.neg, "hg38.srna.neg.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.srna.pos.upstream.flank, "hg38.srna.pos.upstream.flank.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.srna.neg.upstream.flank, "hg38.srna.neg.upstream.flank.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.srna.pos.downstream.flank, "hg38.srna.pos.downstream.flank.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.srna.neg.downstream.flank, "hg38.srna.neg.downstream.flank.bed", sep = "\t", quote = F, col.names = F, row.names = F)


```

use the same binning for sRNA (confirmed by Alastair)


```{bash, chipseq_bedtools_makewindows.sh}

#tile the bed files 
bedtools makewindows -b hg38.srna.pos.bed -n 60 -i winnum > hg38.pos.windows.60w.bed
bedtools makewindows -b hg38.srna.neg.bed -n 60 -i winnum -reverse > hg38.neg.windows.60w.bed


#for upstream flanked
bedtools makewindows -b hg38.srna.pos.upstream.flank.bed -n 20 -i winnum > hg38.pos.upstream.windows.20w.bed
bedtools makewindows -b hg38.srna.neg.upstream.flank.bed -n 20 -i winnum -reverse > hg38.neg.upstream.windows.20w.bed

#for downstream flanked

bedtools makewindows -b hg38.srna.pos.downstream.flank.bed -n 20 -i winnum > hg38.pos.downstream.windows.20w.bed
bedtools makewindows -b hg38.srna.neg.downstream.flank.bed -n 20 -i winnum -reverse > hg38.neg.downstream.windows.20w.bed
# Combine the files as in the Star protocol (instead of R use awk so I can sort)

# Step 1: Combine gene tiles and add 20 to bin numbers (V4)
awk 'BEGIN{OFS="\t"} { $4 = $4 + 20; print }' hg38.pos.windows.60w.bed > srna.tmp
awk 'BEGIN{OFS="\t"} { $4 = $4 + 20; print }' hg38.neg.windows.60w.bed >> srna.tmp

# Step 2: Combine upstream flanks without change
cat hg38.pos.upstream.windows.20w.bed hg38.neg.upstream.windows.20w.bed > upstream.tmp


# Step 3: Combine downstream flanks and add 80 to bin numbers (V4)
awk 'BEGIN{OFS="\t"} { $4 = $4 + 80; print }' hg38.pos.downstream.windows.20w.bed > downstream.tmp
awk 'BEGIN{OFS="\t"} { $4 = $4 + 80; print }' hg38.neg.downstream.windows.20w.bed >> downstream.tmp

# Step 4: Combine all tiles
cat srna.tmp upstream.tmp downstream.tmp > hg38.srna.and.flanks.tiled.bed

# Step 5: Cleanup temporary files
rm srna.tmp upstream.tmp downstream.tmp

# Sort
sort -k1,1 -k2,2n hg38.srna.and.flanks.tiled.bed > hg38.srna.and.flanks.tiled.sorted.bed

```


read in all the remapped bed files with binned coverage
```{r}
avg_bed_pattern = "\\merged.clean.cov_redone.bed"

#function to read all the remapped bed, walk (same as map but does not return a list), then rename the chrmosome, then recalcuklate the distance so it is 5kb 
list.files(pattern = avg_bed_pattern) %>%
  walk(function(x) {
    print(x)
    
    df <- read.table(x, header = FALSE) %>%
      dplyr::rename("chr" = V1, "start" = V2, "end" = V3, "bin" = V4, "COV" = V5) 
   
    # Extract prefix of filename (remove extension)
    prefix <- str_remove(basename(x), "\\.[^.]*$")
    
    # Assign individual dataframe to an object named after the prefix
    assign(paste0(prefix, "_5kb"), df, envir = .GlobalEnv)
  })
```

use the same mapping code for the genes metagene (refer to chip-seq_analysis_clean).
```{r}
avg_bed_pattern = "\\merged.clean.cov_srna.bed"

list.files(path = "/Users/png/Desktop/ATV_lab_projects/Degron_RPB1_project_collab/chipseq_metagene", pattern = avg_bed_pattern, full.names = TRUE) %>%
  walk(function(x) {
    print(x)
    
    df <- read.table(x, header = FALSE) %>%
      dplyr::rename("chr" = V1, "start" = V2, "end" = V3, "bin" = V4, "COV" = V5)
   
    prefix <- str_remove(basename(x), "\\.[^.]*$")
    
    assign(paste0(prefix, "_5kb"), df, envir = .GlobalEnv)
  })
```


normalise the coverage to generate the metagene plot
```{r}

#list all the dataframe which I wanted to loop
sRNA_coverage_list <- list(
  D8L4Y_0h = D8L4Y_0h_merged.clean.cov_srna_5kb,
  D8L4Y_90min = D8L4Y_90min_merged.clean.cov_srna_5kb,
  D8L4Y_8h = D8L4Y_8h_merged.clean.cov_srna_5kb,
  Input = INP_0h_merged.clean.cov_srna_5kb
)

#create function to parse thrugh dataframes to calculate mean and sem then normalized all values to the first bin
#this part is where the normalised signal to background comes in
normalize_coverage <- function(df) {
  df %>%
    group_by(bin) %>%
    summarise(
      COV.mean = mean(COV),
      SEM = sd(COV) / sqrt(n()),
      .groups = "drop"
    ) %>%
    dplyr::mutate(
      COV.norm = COV.mean / COV.mean[1], #normalisation to the background ie region upstream of the TSS
      SEM.norm = SEM / COV.mean[1]
    )
}
 

#lapply the function across all dataframes 
sRNA_plot_list <- lapply(sRNA_coverage_list, normalize_coverage)

#combine the individual dataframes into plots
sRNA_combined.plot <- bind_rows(sRNA_plot_list, .id = "Sample")

#make sure the legend is in the right order
sRNA_combined.plot$Sample <- factor(
  sRNA_combined.plot$Sample,
  levels = c("D8L4Y_0h", "D8L4Y_90min", "D8L4Y_8h", "Input")
)

# plot the coverage plot
sRNA_metagene_plot <- ggplot(sRNA_combined.plot, aes(x = bin, y = COV.norm, color = Sample, fill = Sample)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = COV.norm - SEM.norm, ymax = COV.norm + SEM.norm), alpha = 0.3, color = NA) +
  scale_x_continuous(
    breaks = c(1, 20, 40, 60, 80, 100),
    labels = c("1" = "-5kb", "20" = "TSS", "40" = "33%", "60" = "66%", "80" = "TTS", "100" = "+5kb")
  ) +
  ylab("Normalised coverage (signal / background)") +
  theme_bw() +
  scale_color_manual(values = c( "D8L4Y_0h" = "#F8766D",
    "D8L4Y_90min" = "#00BA38",
    "D8L4Y_8h" = "#619CFF",
    "Input" = "darkgrey"
  )) +
  scale_fill_manual(values = c(
     "D8L4Y_0h" = "#F8766D",
    "D8L4Y_90min" = "#00BA38",
    "D8L4Y_8h" = "#619CFF",
    "Input" = "darkgrey"
  )) +
  theme(
    aspect.ratio = 1,
    axis.text.x = element_text(size = 11, hjust = 0.5),
    axis.text.y = element_text(size = 11),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 12, face = "bold", vjust = 4),
    plot.title = element_text(size = 15, face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )


#supplementary figure 6
sRNA_combined.plot.partial <- sRNA_combined.plot %>% dplyr::filter(Sample != "D8L4Y_8h" )

sRNA_combined.plot.partial$Sample <- factor(
  sRNA_combined.plot.partial$Sample,
  levels = c("D8L4Y_0h", "D8L4Y_90min","Input")
)


#metagene plot 90min vs 0h only

sRNA_metagene_plot.partial <- ggplot(sRNA_combined.plot.partial, aes(x = bin, y = COV.norm, color = Sample, fill = Sample)) +
  geom_line(size = 0.5) +
  geom_ribbon(aes(ymin = COV.norm - SEM.norm, ymax = COV.norm + SEM.norm), alpha = 0.3, color = NA) +
  scale_x_continuous(
    breaks = c(1, 20, 40, 60, 80, 100),
    labels = c("1" = "-5kb", "20" = "TSS", "40" = "33%", "60" = "66%", "80" = "TTS", "100" = "+5kb")
  ) +
  ylab("Normalised coverage (signal/background") +
  theme_bw() +
  scale_color_manual(values = c( "D8L4Y_0h" = "#F8766D",
    "D8L4Y_90min" = "#00BA38",
    "D8L4Y_8h" = "#619CFF",
    "Input" = "darkgrey"
  )) +
  scale_color_manual(values = c( "D8L4Y_0h" = "gray24",
    "D8L4Y_90min" = "dodgerblue3",
    "Input" = "darkgrey"
  )) +
  scale_fill_manual(values = c(
     "D8L4Y_0h" = "gray24",
    "D8L4Y_90min" = "dodgerblue3",
    "Input" = "darkgrey"
  )) +
  theme(text = element_text(family = "sans"),
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.text.x = element_text(size = 14, color = "black"),
    axis.text.y = element_text(size = 14, color = "black"),
     axis.title.x = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    strip.text.x = element_text(size = 14),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    panel.background = element_rect(fill = "white", colour = NA)
  ) +
  ggdag::remove_grid()



```