---
title: "sRNA and lncRNA TT-seq metagene"
author: "Pei Qin (Sabrina) Ng"
date: "2026-07-31"
output: html_document
---


---
lncRNA metagene
---

```{r}
#generate non-overlapping gene annotation (using Granges, rtracklayer , and genomicfeatures)
gtf.dat <- import("/Users/png/Downloads/Homo_sapiens.GRCh38.113.gtf")
lncRNA.gr <- gtf.dat[gtf.dat$type  %in% "gene",]
lncRNA.gr$gene.width <- GenomicRanges::width(lncRNA.gr)
lncRNA.gr <- lncRNA.gr[lncRNA.gr$gene_biotype %in% "lncRNA"]

lncRNA.gr <- keepSeqlevels(lncRNA.gr,c(as.character(1:22),"X","Y"),pruning.mode="coarse")
#lncRNA.gr<- lncRNA.gr[width(lncRNA.gr) >1000]
#convert to dataframe from this step 9Granges can be problematic)
hg38.lncRNA <- as.data.frame(lncRNA.gr) %>% dplyr::rename("chr" = "seqnames") %>% mutate(chr = paste0("chr", chr))


hg38.lncrna.all <- as.data.frame(hg38.lncRNA) %>% 
              mutate(chr = paste0("chr", chr, "_hg38")) %>%
              mutate(start = start - 1) %>%
              mutate(length=end-start) 
#16,917 (> 1kb), 17,268 (all gene length)


# Select pos/neg strand genes; define flanking regions 
#middle region of the metagene plot (redo using filtered bed)
hg38.lncrna.bed.pos.mid.windows.all <- hg38.lncrna.all  %>% 
                                  filter(strand=="+") %>% 
                                  dplyr::select(chr, start, end, length)

hg38.lncrna.bed.neg.mid.windows.all <- hg38.lncrna.all  %>% 
                                  filter(strand=="-") %>% 
                                  dplyr::select(chr, start, end, length)

#flanking regions of the metagene plot (pre-TSS)
hg38.lncrna.bed.pos.early.windows.all <- hg38.lncrna.all %>% 
                                    filter(strand=="+") %>% 
                                    mutate(start=start-5000,end=start+5000) %>% 
                                    dplyr::select(chr, start, end, length) %>% 
                                    filter(start>0)


hg38.lncrna.bed.neg.early.windows.all <- hg38.lncrna.all  %>% 
                                    filter(strand=="-") %>%
                                    mutate(start=start-5000,end=start+5000) %>%
                                    dplyr::select(chr, start, end, length) %>% 
                                    filter(start>0)

#flanking regions of the metagene plot (post-TES)
hg38.lncrna.bed.pos.late.windows.all <- hg38.lncrna.all  %>% 
                                   filter(strand=="+") %>% 
                                   mutate(start=end,end=start+5000) %>% 
                                   dplyr::select(chr, start, end, length) 


hg38.lncrna.bed.neg.late.windows.all <- hg38.lncrna.all  %>% 
                                   filter(strand=="-") %>% 
                                   mutate(start=end,end=start+5000) %>% 
                                   dplyr::select(chr, start, end, length)


# Save files

write.table(hg38.lncrna.bed.pos.mid.windows.all, "hg38.lncrna.pos.mid.windows.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.lncrna.bed.neg.mid.windows.all, "hg38.lncrna.neg.mid.windows.bed", sep = "\t", quote = F, col.names = F, row.names = F)

write.table(hg38.lncrna.bed.pos.early.windows.all, "hg38.lncrna.pos.early.windows.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.lncrna.bed.neg.early.windows.all, "hg38.lncrna.neg.early.windows.bed", sep = "\t", quote = F, col.names = F, row.names = F)

write.table(hg38.lncrna.bed.pos.late.windows.all, "hg38.lncrna.pos.late.windows.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.lncrna.bed.neg.late.windows.all, "hg38.lncrna.neg.late.windows.bed", sep = "\t", quote = F, col.names = F, row.names = F)

```

use the same window splitting as in the gene metagene code then readjust the window numbers
```{r}

#bedtools makewindows -b hg38.lncrna.pos.mid.windows.bed -n 60 -i winnum > hg38.lncrna.pos.mid.windows.60w.bed
#bedtools makewindows -b hg38.lncrna.neg.mid.windows.bed -n 60 -i winnum -reverse > hg38.lncrna.neg.mid.windows.60w.bed

#bedtools makewindows -b hg38.lncrna.pos.early.windows.bed -n 20 -i winnum > hg38.lncrna.pos.early.windows.20w.bed
#bedtools makewindows -b hg38.lncrna.neg.early.windows.bed -n 20 -i winnum -reverse > hg38.lncrna.neg.early.windows.20w.bed

#bedtools makewindows -b hg38.lncrna.pos.late.windows.bed -n 20 -i winnum > hg38.lncrna.pos.late.windows.20w.bed
#bedtools makewindows -b hg38.lncrna.neg.late.windows.bed -n 20 -i winnum -reverse > hg38.lncrna.neg.late.windows.20w.bed
lncrna_TT_metagene_path <- "/Users/png/Desktop/ATV_lab_projects/Degron_RPB1_project_collab/TTseq_metagene/lncrna"


hg38.lncrna.bed.pos.mid.windows.all <- read.table(
  file.path(lncrna_TT_metagene_path, "hg38.lncrna.pos.mid.windows.60w.bed"),
  header = FALSE
) %>%
  mutate(V4 = V4 + 20)

hg38.lncrna.bed.neg.mid.windows.all <-  read.table(
  file.path(lncrna_TT_metagene_path, "hg38.lncrna.neg.mid.windows.60w.bed"),
  header = FALSE
) %>%
  mutate(V4 = V4 + 20) 

hg38.lncrna.bed.pos.early.windows.all <- read.table(file.path(lncrna_TT_metagene_path, "hg38.lncrna.pos.early.windows.20w.bed"), header=FALSE)   

hg38.lncrna.bed.neg.early.windows.all <- read.table(file.path(lncrna_TT_metagene_path,"hg38.lncrna.neg.early.windows.20w.bed"),header=FALSE) %>% 
                                    mutate(V4=V4+80)  

hg38.lncrna.bed.pos.late.windows.all <- read.table(file.path(lncrna_TT_metagene_path,"hg38.lncrna.pos.late.windows.20w.bed"),header=FALSE) %>%
                                   mutate(V4=V4+80) 

hg38.lncrna.bed.neg.late.windows.all <- read.table(file.path(lncrna_TT_metagene_path,"hg38.lncrna.neg.late.windows.20w.bed"),header=FALSE) 

# Combine the corrected bed files 
hg38.lncrna.corrected.all.bed <- rbind(hg38.lncrna.bed.pos.early.windows.all, hg38.lncrna.bed.pos.mid.windows.all, hg38.lncrna.bed.pos.late.windows.all,hg38.lncrna.bed.neg.early.windows.all, hg38.lncrna.bed.neg.mid.windows.all, hg38.lncrna.bed.neg.late.windows.all)
hg38.lncrna.pos.corrected.all.bed <- rbind(hg38.lncrna.bed.pos.early.windows.all, hg38.lncrna.bed.pos.mid.windows.all, hg38.lncrna.bed.pos.late.windows.all)
hg38.lncrna.neg.corrected.all.bed <- rbind(hg38.lncrna.bed.neg.early.windows.all, hg38.lncrna.bed.neg.mid.windows.all, hg38.lncrna.bed.neg.late.windows.all)

# Save files
write.table(hg38.lncrna.corrected.all.bed, "hg38.lncrna.corrected.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.lncrna.pos.corrected.all.bed, "hg38.lncrna.pos.corrected.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.lncrna.neg.corrected.all.bed, "hg38.lncrna.neg.corrected.bed", sep = "\t", quote = F, col.names = F, row.names = F)
```



plot the metagene plot using spike-in normalisation factor bigwigs
Once completed the mapping of the coverage onto the binned genome regions (in bed format), generate the plot (code from Ivan)


```{r}

### Plot 

# Define sample names and corresponding file prefixes
samples <- c("0h", "90min", "8h")
prefixes <- c("0h", "90min", "8h")

# Initialize empty list to store plot data frames
plot_list_lncrna_TT <- list()

for(i in seq_along(samples)) {
  sample_name <- samples[i]
  prefix <- prefixes[i]
  
  # Read forward and reverse coverage files and combine
  coverage <- rbind(
    read.table(file.path(lncrna_TT_metagene_path, paste0(prefix, "_spikein.norm._merged_hg38_fwd.lncrna.cov.bed")), header = FALSE),
    read.table(file.path(lncrna_TT_metagene_path, paste0(prefix, "_spikein.norm._merged_hg38_rev.lncrna.cov.bed")), header = FALSE)
  )
  
  # Rename columns
  colnames(coverage) <- c("chr", "start", "end", "bin", "COV")
  
  # Add dist column
  coverage <- coverage %>%
    mutate(dist = (as.numeric(bin) - 1) * 0.0505050505)
  
  # Summarise mean and SEM by dist
  plot_df_lncrna <- coverage %>%
    group_by(dist) %>%
    summarise(
      COV.mean = mean(COV),
      SEM = sd(COV)/sqrt(n())
    ) %>%
    mutate(Condition = case_when(
      sample_name == "0h" ~ "0h",
      sample_name == "90min" ~ "90min",
      sample_name == "8h" ~ "8h"
    ))
  
  # Store result in list
  plot_list_lncrna_TT[[sample_name]] <- plot_df_lncrna
}

# Combine all plot data frames
Total_lncrna.plot.TT <- bind_rows(plot_list_lncrna_TT)
#only 0h and 90mins

Total_lncrna.plot.TT_part <- Total_lncrna.plot.TT %>% filter(Condition != "8h")
# Formatting for better aesthetics

Total_lncrna.plot.TT$Condition <- factor(Total_lncrna.plot.TT$Condition, 
                                     levels = c("0h", "90min", "8h"))

Total_lncrna.plot.TT_part$Condition <- factor(Total_lncrna.plot.TT_part$Condition, 
                                     levels = c("0h", "90min"))
#Plot the metagene plot
Total_lncrna.plot.TT_p <- Total_lncrna.plot.TT %>%
  ggplot(aes(x = dist, y = COV.mean, color = Condition, fill = Condition)) +
  geom_line(size = 1) +
  scale_color_manual(values = c( "0h" = "gray24",
    "90min" = "dodgerblue3",
    "8h" = "darkgreen",
    "Input" = "darkgrey"
  )) +
  scale_fill_manual(values = c(
     "0h" = "gray24",
    "90min" = "dodgerblue3",
    "8h" = "darkgreen",
    "Input" = "darkgrey"
  )) +
  geom_ribbon(aes(ymin = COV.mean - SEM, ymax = COV.mean + SEM), alpha = 0.2, color = NA) +
  #scale_color_manual(values = colors) +
  #scale_fill_manual(values = colors) +
  #scale_linetype_manual(values = line_types) + # Add this to set linetypes
  scale_x_continuous(breaks = seq(0, 5, by = 1), 
                     labels = c("0" = "-5kb", "1" = "TSS", "2" = "33%", "3" = "66%", "4" = "TTS", "5" = "+5kb")) +
  guides(linetype = guide_legend(override.aes = list(fill = NA))) + # Adjust the legend for linetypes
  xlab("Position") + ylab("Spike-in normalised coverage") + 
  ggtitle("") +
  theme_bw() +
  theme(
    #aspect.ratio = 1,
    legend.title = element_text(size = 14, face = "bold"), 
    legend.text = element_text(size = 10),
    axis.text.x = element_text(size = 11, angle = 0, vjust = 0, hjust = 0.5),
    axis.text.y = element_text(size = 11),
    axis.title.x = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title.y = element_text(size = 14, face = "bold", vjust = 2),
    plot.title = element_text(size = 16, face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    legend.background = element_blank(),
    legend.key = element_blank()
  ) +
  guides(fill = "none") + ggtitle("TT-seq metagene plot for lncRNAs (spike-in normalised)") # Hide the fill legend


#plot 90mins and 0h for lncrna
Total_lncrna.plot.TT_part_plot <- Total_lncrna.plot.TT_part %>%
  ggplot(aes(x = dist, y = COV.mean, color = Condition, fill = Condition)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = COV.mean - SEM, ymax = COV.mean + SEM), alpha = 0.2, color = NA) +
  #scale_color_manual(values = colors) +
  #scale_fill_manual(values = colors) +
  #scale_linetype_manual(values = line_types) + # Add this to set linetypes
  scale_x_continuous(breaks = seq(0, 5, by = 1), 
                     labels = c("0" = "-5kb", "1" = "TSS", "2" = "33%", "3" = "66%", "4" = "TTS", "5" = "+5kb")) +
  guides(linetype = guide_legend(override.aes = list(fill = NA))) + # Adjust the legend for linetypes
  xlab("Position") + ylab("Spike-in normalised coverage") + 
  ggtitle("") +
  scale_color_manual(values = c( "0h" = "gray24",
    "90min" = "dodgerblue3",
    "Input" = "darkgrey"
  )) +
  scale_fill_manual(values = c(
     "0h" = "gray24",
    "90min" = "dodgerblue3",
    "Input" = "darkgrey"
  )) +
 theme(text = element_text(family = "sans"),
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.text.x = element_text(size = 14, color = "black"),
    axis.text.y = element_text(size = 14, color = "black"),
     axis.title.x = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    strip.text.x = element_text(size = 14),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    panel.background = element_rect(fill = "white", colour = NA)
  ) 

```



---
sRNA metagene
---

Generate non-overlapping annotations for make windows (all genes)

```{r}
library(GenomicRanges)
library(rtracklayer)
gtf.dat <- import("/Users/png/Downloads/Homo_sapiens.GRCh38.113.gtf")
srna.gr <- gtf.dat[gtf.dat$type  %in% "gene",]
srna.gr$gene.width <- GenomicRanges::width(srna.gr)
srna.gr<- srna.gr[srna.gr$gene_biotype %in% "snRNA"]
srna.gr <- keepSeqlevels(srna.gr,c(as.character(1:22),"X","Y"),pruning.mode="coarse")
srna.gr <- srna.gr[width(srna.gr ) >60]
#convert to dataframe from this step 9Granges can be problematic)
hg38.srna <- as.data.frame(srna.gr) %>% dplyr::rename("chr" = "seqnames") %>% mutate(chr = paste0("chr", chr))


#convert to dataframe from this step 9Granges can be problematic)
hg38.srna.all <- as.data.frame(hg38.srna) %>% 
              mutate(chr = paste0("chr", chr, "_hg38")) %>%
              mutate(start = start - 1) %>%
              mutate(length=end-start) 
#16,917 (> 1kb), 17,268 (all gene length)


# Select pos/neg strand genes; define flanking regions 
#middle region of the metagene plot (redo using filtered bed)
hg38.srna.bed.pos.mid.windows.all <- hg38.srna.all  %>% 
                                  filter(strand=="+") %>% 
                                  dplyr::select(chr, start, end, length)

hg38.srna.bed.neg.mid.windows.all <- hg38.srna.all  %>% 
                                  filter(strand=="-") %>% 
                                  dplyr::select(chr, start, end, length)

#flanking regions of the metagene plot (pre-TSS)
hg38.srna.bed.pos.early.windows.all <- hg38.srna.all %>% 
                                    filter(strand=="+") %>% 
                                    mutate(start=start-5000,end=start+5000) %>% 
                                    dplyr::select(chr, start, end, length) %>% 
                                    filter(start>0)


hg38.srna.bed.neg.early.windows.all <- hg38.srna.all  %>% 
                                    filter(strand=="-") %>%
                                    mutate(start=start-5000,end=start+5000) %>%
                                    dplyr::select(chr, start, end, length) %>% 
                                    filter(start>0)

#flanking regions of the metagene plot (post-TES)
hg38.srna.bed.pos.late.windows.all <- hg38.srna.all  %>% 
                                   filter(strand=="+") %>% 
                                   mutate(start=end,end=start+5000) %>% 
                                   dplyr::select(chr, start, end, length) 


hg38.srna.bed.neg.late.windows.all <- hg38.srna.all  %>% 
                                   filter(strand=="-") %>% 
                                   mutate(start=end,end=start+5000) %>% 
                                   dplyr::select(chr, start, end, length)


# Save files

write.table(hg38.srna.bed.pos.mid.windows.all, "hg38.srna.pos.mid.windows.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.srna.bed.neg.mid.windows.all, "hg38.srna.neg.mid.windows.bed", sep = "\t", quote = F, col.names = F, row.names = F)

write.table(hg38.srna.bed.pos.early.windows.all, "hg38.srna.pos.early.windows.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.srna.bed.neg.early.windows.all, "hg38.srna.neg.early.windows.bed", sep = "\t", quote = F, col.names = F, row.names = F)

write.table(hg38.srna.bed.pos.late.windows.all, "hg38.srna.pos.late.windows.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.srna.bed.neg.late.windows.all, "hg38.srna.neg.late.windows.bed", sep = "\t", quote = F, col.names = F, row.names = F)

 
```


use the same window splitting as in the gene metagene code then readjust the window numbers
```{r}

#bedtools makewindows -b hg38.srna.pos.mid.windows.bed -n 60 -i winnum > hg38.srna.pos.mid.windows.60w.bed
#bedtools makewindows -b hg38.srna.neg.mid.windows.bed -n 60 -i winnum -reverse > hg38.srna.neg.mid.windows.60w.bed

#bedtools makewindows -b hg38.srna.pos.early.windows.bed -n 20 -i winnum > hg38.srna.pos.early.windows.20w.bed
#bedtools makewindows -b hg38.srna.neg.early.windows.bed -n 20 -i winnum -reverse > hg38.srna.neg.early.windows.20w.bed

#bedtools makewindows -b hg38.srna.pos.late.windows.bed -n 20 -i winnum > hg38.srna.pos.late.windows.20w.bed
#bedtools makewindows -b hg38.srna.neg.late.windows.bed -n 20 -i winnum -reverse > hg38.srna.neg.late.windows.20w.bed
sRNA_TT_metagene_path <- "/Users/png/Desktop/ATV_lab_projects/Degron_RPB1_project_collab/TTseq_metagene/srna"


hg38.srna.bed.pos.mid.windows.all <- read.table(
  file.path(sRNA_TT_metagene_path, "hg38.srna.pos.mid.windows.60w.bed"),
  header = FALSE
) %>%
  mutate(V4 = V4 + 20)

hg38.srna.bed.neg.mid.windows.all <-  read.table(
  file.path(sRNA_TT_metagene_path, "hg38.srna.neg.mid.windows.60w.bed"),
  header = FALSE
) %>%
  mutate(V4 = V4 + 20) 

hg38.srna.bed.pos.early.windows.all <- read.table(file.path(sRNA_TT_metagene_path, "hg38.srna.pos.early.windows.20w.bed"), header=FALSE)   

hg38.srna.bed.neg.early.windows.all <- read.table(file.path(sRNA_TT_metagene_path,"hg38.srna.neg.early.windows.20w.bed"),header=FALSE) %>% 
                                    mutate(V4=V4+80)  

hg38.srna.bed.pos.late.windows.all <- read.table(file.path(sRNA_TT_metagene_path,"hg38.srna.pos.late.windows.20w.bed"),header=FALSE) %>%
                                   mutate(V4=V4+80) 

hg38.srna.bed.neg.late.windows.all <- read.table(file.path(sRNA_TT_metagene_path,"hg38.srna.neg.late.windows.20w.bed"),header=FALSE) 

# Combine the corrected bed files 
hg38.srna.corrected.all.bed <- rbind(hg38.srna.bed.pos.early.windows.all, hg38.srna.bed.pos.mid.windows.all, hg38.srna.bed.pos.late.windows.all,hg38.srna.bed.neg.early.windows.all, hg38.srna.bed.neg.mid.windows.all, hg38.srna.bed.neg.late.windows.all)
hg38.srna.pos.corrected.all.bed <- rbind(hg38.srna.bed.pos.early.windows.all, hg38.srna.bed.pos.mid.windows.all, hg38.srna.bed.pos.late.windows.all)
hg38.srna.neg.corrected.all.bed <- rbind(hg38.srna.bed.neg.early.windows.all, hg38.srna.bed.neg.mid.windows.all, hg38.srna.bed.neg.late.windows.all)

# Save files
write.table(hg38.srna.corrected.all.bed, "hg38.srna.corrected.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.srna.pos.corrected.all.bed, "hg38.srna.pos.corrected.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.srna.neg.corrected.all.bed, "hg38.srna.neg.corrected.bed", sep = "\t", quote = F, col.names = F, row.names = F)
```


plot the metagene plot using spike-in normalisation factor bigwigs
Once completed the mapping of the coverage onto the binned genome regions (in bed format), generate the plot (code from Ivan)


```{r}


### Plot 

# Define sample names and corresponding file prefixes
samples <- c("0h", "90min", "8h")
prefixes <- c("0h", "90min", "8h")

# Initialize empty list to store plot data frames
plot_list_srna_TT <- list()

for(i in seq_along(samples)) {
  sample_name <- samples[i]
  prefix <- prefixes[i]
  
  # Read forward and reverse coverage files and combine
  coverage <- rbind(
    read.table(file.path(sRNA_TT_metagene_path, paste0(prefix, "_spikein.norm._merged_hg38_fwd.srna2.cov.bed")), header = FALSE),
    read.table(file.path(sRNA_TT_metagene_path, paste0(prefix, "_spikein.norm._merged_hg38_rev.srna2.cov.bed")), header = FALSE)
  )
  
  # Rename columns
  colnames(coverage) <- c("chr", "start", "end", "bin", "COV")
  
  # Add dist column
  coverage <- coverage %>%
    mutate(dist = (as.numeric(bin) - 1) * 0.0505050505)
  
  # Summarise mean and SEM by dist
  plot_df_srna <- coverage %>%
    group_by(dist) %>%
    summarise(
      COV.mean = mean(COV),
      SEM = sd(COV)/sqrt(n())
    ) %>%
    mutate(Condition = case_when(
      sample_name == "0h" ~ "0h",
      sample_name == "90min" ~ "90min",
      sample_name == "8h" ~ "8h"
    ))
  
  # Store result in list
  plot_list_srna_TT[[sample_name]] <- plot_df_srna
}

# Combine all plot data frames
Total_srna.plot.TT <- bind_rows(plot_list_srna_TT)
#only 0h and 90mins

Total_srna.plot.TT_part <- Total_srna.plot.TT %>% filter(Condition != "8h")
# Formatting for better aesthetics

Total_srna.plot.TT$Condition <- factor(Total_srna.plot.TT$Condition, 
                                     levels = c("0h", "90min", "8h"))

Total_srna.plot.TT_part$Condition <- factor(Total_srna.plot.TT_part$Condition, 
                                     levels = c("0h", "90min"))
#Plot the metagene plot
Total_srna.plot.TT_p <- Total_srna.plot.TT %>%
  ggplot(aes(x = dist, y = COV.mean, color = Condition, fill = Condition)) +
  geom_line(size = 1) +
  scale_color_manual(values = c( "0h" = "gray24",
    "90min" = "dodgerblue3",
    "8h" = "darkgreen",
    "Input" = "darkgrey"
  )) +
  scale_fill_manual(values = c(
     "0h" = "gray24",
    "90min" = "dodgerblue3",
    "8h" = "darkgreen",
    "Input" = "darkgrey"
  )) +
  geom_ribbon(aes(ymin = COV.mean - SEM, ymax = COV.mean + SEM), alpha = 0.2, color = NA) +
  #scale_color_manual(values = colors) +
  #scale_fill_manual(values = colors) +
  #scale_linetype_manual(values = line_types) + # Add this to set linetypes
  scale_x_continuous(breaks = seq(0, 5, by = 1), 
                     labels = c("0" = "-5kb", "1" = "TSS", "2" = "33%", "3" = "66%", "4" = "TTS", "5" = "+5kb")) +
  scale_y_continuous(limits = c(0,400))+
  guides(linetype = guide_legend(override.aes = list(fill = NA))) + # Adjust the legend for linetypes
  xlab("Position") + ylab("Spike-in normalised coverage") + 
  ggtitle("") +
  theme_bw() +
  theme(
    #aspect.ratio = 1,
    legend.title = element_text(size = 14, face = "bold"), 
    legend.text = element_text(size = 10),
    axis.text.x = element_text(size = 11, angle = 0, vjust = 0, hjust = 0.5),
    axis.text.y = element_text(size = 11),
    axis.title.x = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title.y = element_text(size = 14, face = "bold", vjust = 2),
    plot.title = element_text(size = 16, face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    legend.background = element_blank(),
    legend.key = element_blank()
  ) +
  guides(fill = "none") + ggtitle("TT-seq metagene plot for sRNAs (spike-in normalised)") # Hide the fill legend


#plot 90mins and 0h for sRNA
Total_srna.plot.TT_part_plot <- Total_srna.plot.TT_part %>%
  ggplot(aes(x = dist, y = COV.mean, color = Condition, fill = Condition)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = COV.mean - SEM, ymax = COV.mean + SEM), alpha = 0.2, color = NA) +
  #scale_color_manual(values = colors) +
  #scale_fill_manual(values = colors) +
  #scale_linetype_manual(values = line_types) + # Add this to set linetypes
  scale_x_continuous(breaks = seq(0, 5, by = 1), 
                     labels = c("0" = "-5kb", "1" = "TSS", "2" = "33%", "3" = "66%", "4" = "TTS", "5" = "+5kb")) +
  scale_y_continuous(limits = c(0,200))+
  guides(linetype = guide_legend(override.aes = list(fill = NA))) + # Adjust the legend for linetypes
  xlab("Position") + ylab("Spike-in normalised coverage") + 
  ggtitle("") +
  scale_color_manual(values = c( "0h" = "gray24",
    "90min" = "dodgerblue3",
    "Input" = "darkgrey"
  )) +
  scale_fill_manual(values = c(
     "0h" = "gray24",
    "90min" = "dodgerblue3",
    "Input" = "darkgrey"
  )) +
 theme(text = element_text(family = "sans"),
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.text.x = element_text(size = 14, color = "black"),
    axis.text.y = element_text(size = 14, color = "black"),
     axis.title.x = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    strip.text.x = element_text(size = 14),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    panel.background = element_rect(fill = "white", colour = NA)
  ) 

```


---
combine protein coding, sRNA and lncRNA metagene plots together
---

```{r}
library(patchwork)

#all three conditions
Total_srna.plot.TT_p+TT_metagene_si_p+Total_lncrna.plot.TT_p 


#90mins and 0h ones only
Total_srna.plot.TT_part_plot +TT_metagene_si_p_part + Total_lncrna.plot.TT_part_plot
```

