---
title: "TT-seq"
output: html_document
date: "2025-07-18"
---
read the required counts data

```{r}
#load required libraries
library(tidyverse)
library(readr)
library(dplyr)
library(ggpubr)
library(purrr)
library(stringr)
#read in all summary count files 
count_data_pattern = "\\.txt"


Degron_count_data <- list.files(path = "/Users/png/Desktop/ATV_lab_projects/Degron_RPB1_project_collab/counts/",
                         pattern = count_data_pattern, full.names = TRUE) %>%
    map(function(x) {print(x)
        readr::read_table(x, col_names = F, skip =2)}
        ) %>% purrr::reduce(right_join,c("X1","X2", "X3","X4", "X5", "X6")) %>% 
    dplyr::rename(gene_id= "X1", chrom = "X2", start = "X3", end = "X4", strand = "X5", length = "X6")  %>%
    rename_with(~str_replace(list.files(path = "/Users/png/Desktop/ATV_lab_projects/Degron_RPB1_project_collab/counts/",
                         pattern = count_data_pattern) , "_ensembl_featureCounts\\..*", ""), 7:15)

```

Perform Deseq2 normalisation for the dataset. 
```{r}
library(DESeq2)
library(tidyverse)
#clean up the counts files
Degron_counts_df <- Degron_count_data %>%
  mutate(gene_id = ifelse(grepl("_sacCer3", chrom), paste0(gene_id, "_sacCer3"), gene_id)) %>%
    dplyr::select('gene_id',
           7:15) %>% 
          column_to_rownames('gene_id') 

#generate metadata
Degron_metadata <- data.frame(sample = colnames(Degron_counts_df[1:9]), 
                         group = c(rep("0h", 3), rep("8h", 3), rep("90min", 3)) )    
#make sure group is a factor
Degron_metadata$group<-as.factor(Degron_metadata$group)
```

Read in counts as Deseq object
```{r}
dds_Degron <- DESeqDataSetFromMatrix(countData=Degron_counts_df, 
                              colData=Degron_metadata, 
                              design= ~group)


#dds$group <- relevel(dds$group, ref = "ctrl")
#dds$group <- factor(dds$group, levels = c("ctrl","VEGF", "VEGFDMFu", "DMFu"))
#standard workflow ie estimate dispersion, estimate size factor and wald test for DE

# Identify the spike-in genes subsets
control_genes_Degron <- grepl("sacCer3$", rownames(dds_Degron))
# Estimate size factors (factoring in the spike in)
dds_Degron <- estimateSizeFactors(dds_Degron, controlGenes = control_genes_Degron)

#export size factor to be used in normalising reads for coverage computation

Degron_sizeFactors <- dds_Degron$sizeFactor %>% as.data.frame() %>% rownames_to_column()
write_csv(Degron_sizeFactors, "Degron_TTseq.Deseq2.sizefactors.csv", col_names = FALSE)

#if need to do filtering can do after estimate sizafactor
#for DEG between KO and WT, I filtered away the ones with low counts
keep_Degron <- rowSums(counts(dds_Degron)) >= 10
dds_Degron <- dds_Degron[keep_Degron,]
```


Remove the spike-in yeast genes from the data before the next step
```{r}
#remove the spike-ins from the dataset before continue-ing
exclude_genes_Degron <- grepl("sacCer3$", rownames(dds_Degron))

#subset the DESeqDataSet object to exclude these genes
dds_filtered_Degron <- dds_Degron[!exclude_genes_Degron,]

#quick check of the genes being removed from dds
#rownames(dds)[exclude_genes]
```


Plot PCA plots for all samples

```{r}
#specify the "group" variable for colData(dds) as the variable for intgroup
intgroup_Degron <- "group"

# Transform the data using rlog transformation (using only genes from hg38)
rlog_Degron <- rlogTransformation(dds_filtered_Degron)



# Create PCA plot (all samples together)
Degron_PCA_plot <-  plotPCA(rlog_Degron, intgroup = intgroup_Degron) + 
             theme_bw() +
             theme(
             axis.title = element_text(size = 14),  # Adjust the font size for axis titles
             axis.text = element_text(size = 14),   # Adjust the font size for axis text
             legend.title = element_text(size = 14),# Adjust the font size for legend title
             legend.text = element_text(size = 14), 
             aspect.ratio = 1  # Adjust the font size for legend text
            ) + ggtitle("Degron_PCA plot combined using only top n =500 variance")

#using ntop=500 top features by variance (only hg38 genes)

#separate data by ELOA KO and Wt
# Perform PCA on rlog-transformed data
Degron_pca_result <- prcomp(t(assay(rlog_Degron)))  # Transpose to get samples as rows

# Get PCA coordinates (PC1, PC2, etc.)
Degron_pca_data <- as.data.frame(Degron_pca_result$x)

# Create a new metadata column based on sample names
Degron_pca_data$Sample <- colnames(rlog_Degron)  # Sample names
Degron_pca_data$Group <- ifelse(grepl("^0h_", Degron_pca_data$Sample), 
                                "0h", 
                                ifelse(grepl("^8h_", Degron_pca_data$Sample), 
                                       "8h", 
                                       "90min"))
Degron_pca_data$Group <- factor(Degron_pca_data$Group, levels = c("0h", "90min", "8h"))
# Calculate the variance explained by each principal component
Degron_var_explained <- summary(Degron_pca_result)$importance[2, ] * 100 


Degron_pca_data$Group <- factor(Degron_pca_data$Group, 
                                     levels = c("0h", "90min", "8h"))
# plot PCA (ggplot2)
Degron_PCA_plot_group <- ggplot(Degron_pca_data, aes(x = PC1, y = PC2, colour = Group)) +
  geom_point(size = 3, alpha = 0.5) +
  labs(title = "PCA Plot", x = "PC1", y = "PC2") +
  theme_bw() +
  labs(
    title = "PCA Plot",
    x = paste0("PC1 (", round(Degron_var_explained[1], 2), "%)"),  # PC1 variance explained
    y = paste0("PC2 (", round(Degron_var_explained[2], 2), "%)")   # PC2 variance explained
  ) 


```


---
Metagene plot analysis
---

Generate non-overlapping annotations for make windows. Use only genes > 1000 to minimise noise in the metagene plot.

```{r}
library(GenomicRanges)
library(rtracklayer)
#annotation file (the same as in ELOA P1)
#hg38.gtf <- read.table("/Users/png/Downloads/Homo_sapiens.GRCh38.113.gtf", header=FALSE, sep= "\t")

#generate non-overlapping gene annotation (using Granges, rtracklayer , and genomicfeatures)
gtf.dat <- import("/Users/png/Downloads/Homo_sapiens.GRCh38.113.gtf")
gene.gr <- gtf.dat[gtf.dat$type %in% "gene",]
names(gene.gr) <- gene.gr$gene_id
gene.gr$gene.width <- GenomicRanges::width(gene.gr)
gene.gr$gene_name_column <- gene.gr$gene_name

gene.gr <- gene.gr[width(gene.gr) >1000]
gene.gr <- gene.gr[gene.gr$gene_biotype %in% "protein_coding"]
gene.gr <- gene.gr[countOverlaps(gene.gr,gene.gr) == 1]
gene.gr <- keepSeqlevels(gene.gr,c(as.character(1:22),"X","Y"),pruning.mode="coarse")
#convert to dataframe from this step 9Granges can be problematic)
hg38.genes <- as.data.frame(gene.gr) %>% 
              dplyr::rename("chr" = "seqnames") %>% 
              mutate(chr = paste0("chr", chr, "_hg38")) %>%
              mutate(start = start - 1) %>%
              mutate(length=end-start) 
#16,917


# Select pos/neg strand genes; define flanking regions 
#middle region of the metagene plot (redo using filtered bed)
hg38.genes.bed.pos.mid.windows <- hg38.genes %>% 
                                  filter(strand=="+") %>% 
                                  dplyr::select(chr, start, end, length)

hg38.genes.bed.neg.mid.windows <- hg38.genes %>% 
                                  filter(strand=="-") %>% 
                                  dplyr::select(chr, start, end, length)

#flanking regions of the metagene plot (pre-TSS)
hg38.genes.bed.pos.early.windows <- hg38.genes %>% 
                                    filter(strand=="+") %>% 
                                    mutate(start=start-5000,end=start+5000) %>% 
                                    dplyr::select(chr, start, end, length) %>% 
                                    filter(start>0)


hg38.genes.bed.neg.early.windows <- hg38.genes  %>% 
                                    filter(strand=="-") %>%
                                    mutate(start=start-5000,end=start+5000) %>%
                                    dplyr::select(chr, start, end, length) %>% 
                                    filter(start>0)

#flanking regions of the metagene plot (post-TES)
hg38.genes.bed.pos.late.windows <- hg38.genes  %>% 
                                   filter(strand=="+") %>% 
                                   mutate(start=end,end=start+5000) %>% 
                                   dplyr::select(chr, start, end, length) 


hg38.genes.bed.neg.late.windows <- hg38.genes  %>% 
                                   filter(strand=="-") %>% 
                                   mutate(start=end,end=start+5000) %>% 
                                   dplyr::select(chr, start, end, length)


# Save files

write.table(hg38.genes.bed.pos.mid.windows, "hg38.genes.fil1000.pos.mid.windows.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.genes.bed.neg.mid.windows, "hg38.genes.fil1000.neg.mid.windows.bed", sep = "\t", quote = F, col.names = F, row.names = F)

write.table(hg38.genes.bed.pos.early.windows, "hg38.genes.fil1000.pos.early.windows.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.genes.bed.neg.early.windows, "hg38.genes.fil1000.neg.early.windows.bed", sep = "\t", quote = F, col.names = F, row.names = F)

write.table(hg38.genes.bed.pos.late.windows, "hg38.genes.fil1000.pos.late.windows.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.genes.bed.neg.late.windows, "hg38.genes.fil1000.neg.late.windows.bed", sep = "\t", quote = F, col.names = F, row.names = F)

 
```
Generate non-overlapping annotations for make windows. 

```{r}
library(GenomicRanges)
library(rtracklayer)
#annotation file (the same as in ELOA P1)
#hg38.gtf <- read.table("/Users/png/Downloads/Homo_sapiens.GRCh38.113.gtf", header=FALSE, sep= "\t")

#generate non-overlapping gene annotation (using Granges, rtracklayer , and genomicfeatures)
gtf.dat <- import("/Users/png/Downloads/Homo_sapiens.GRCh38.113.gtf")
gene.gr.all <- gtf.dat[gtf.dat$type %in% "gene",]
names(gene.gr.all) <- gene.gr.all$gene_id
gene.gr.all$gene.width <- GenomicRanges::width(gene.gr.all)
gene.gr.all$gene_name_column <- gene.gr.all$gene_name

#gene.gr <- gene.gr[width(gene.gr) >1000]
gene.gr.all <- gene.gr.all[gene.gr.all$gene_biotype %in% "protein_coding"]
gene.gr.all <- gene.gr.all[countOverlaps(gene.gr.all,gene.gr.all) == 1]
gene.gr.all <- keepSeqlevels(gene.gr.all,c(as.character(1:22),"X","Y"),pruning.mode="coarse")
#convert to dataframe from this step 9Granges can be problematic)
hg38.genes.all <- as.data.frame(gene.gr.all) %>% 
              dplyr::rename("chr" = "seqnames") %>% 
              mutate(chr = paste0("chr", chr, "_hg38")) %>%
              mutate(start = start - 1) %>%
              mutate(length=end-start) 
#16,917 (> 1kb), 17,268 (all gene length)


# Select pos/neg strand genes; define flanking regions 
#middle region of the metagene plot (redo using filtered bed)
hg38.genes.bed.pos.mid.windows.all <- hg38.genes.all %>% 
                                  filter(strand=="+") %>% 
                                  dplyr::select(chr, start, end, length)

hg38.genes.bed.neg.mid.windows.all <- hg38.genes.all %>% 
                                  filter(strand=="-") %>% 
                                  dplyr::select(chr, start, end, length)

#flanking regions of the metagene plot (pre-TSS)
hg38.genes.bed.pos.early.windows.all <- hg38.genes.all %>% 
                                    filter(strand=="+") %>% 
                                    mutate(start=start-5000,end=start+5000) %>% 
                                    dplyr::select(chr, start, end, length) %>% 
                                    filter(start>0)


hg38.genes.bed.neg.early.windows.all <- hg38.genes.all  %>% 
                                    filter(strand=="-") %>%
                                    mutate(start=start-5000,end=start+5000) %>%
                                    dplyr::select(chr, start, end, length) %>% 
                                    filter(start>0)

#flanking regions of the metagene plot (post-TES)
hg38.genes.bed.pos.late.windows.all <- hg38.genes.all  %>% 
                                   filter(strand=="+") %>% 
                                   mutate(start=end,end=start+5000) %>% 
                                   dplyr::select(chr, start, end, length) 


hg38.genes.bed.neg.late.windows.all <- hg38.genes.all  %>% 
                                   filter(strand=="-") %>% 
                                   mutate(start=end,end=start+5000) %>% 
                                   dplyr::select(chr, start, end, length)


# Save files

write.table(hg38.genes.bed.pos.mid.windows.all, "hg38.genes.pos.mid.windows.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.genes.bed.neg.mid.windows.all, "hg38.genes.neg.mid.windows.bed", sep = "\t", quote = F, col.names = F, row.names = F)

write.table(hg38.genes.bed.pos.early.windows, "hg38.genes.pos.early.windows.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.genes.bed.neg.early.windows, "hg38.genes.neg.early.windows.bed", sep = "\t", quote = F, col.names = F, row.names = F)

write.table(hg38.genes.bed.pos.late.windows, "hg38.genes.pos.late.windows.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.genes.bed.neg.late.windows, "hg38.genes.neg.late.windows.bed", sep = "\t", quote = F, col.names = F, row.names = F)

 
```

After splitting the bed files using bedtools makewindow as per the original script (bedtools_makewindows_**.sh). Correct the bed files (so the positives and negatives matches) before mapping it to the blacklist. The mapping to the blacklist-cleaned reads should be done using bash on cluster (bedsort first)



```{bash, bedtools_makewindows.sh}
#tile the bed files 
#bedtools makewindows -b hg38.genes.pos.mid.windows.bed -n 60 -i winnum > hg38.genes.pos.mid.windows.60w.bed
#bedtools makewindows -b hg38.genes.neg.mid.windows.bed -n 60 -i winnum -reverse > hg38.genes.neg.mid.windows.60w.bed

#bedtools makewindows -b hg38.genes.pos.early.windows.bed -n 20 -i winnum > hg38.genes.pos.early.windows.20w.bed
#bedtools makewindows -b hg38.genes.neg.early.windows.bed -n 20 -i winnum -reverse > hg38.genes.neg.early.windows.20w.bed

#bedtools makewindows -b hg38.genes.pos.late.windows.bed -n 20 -i winnum > hg38.genes.pos.late.windows.20w.bed
#bedtools makewindows -b hg38.genes.neg.late.windows.bed -n 20 -i winnum -reverse > hg38.genes.neg.late.windows.20w.bed

#ignore the fail error message which says that there will be one line where the start > end. the code will still generate the output
```


Import and adjust window numbers for metagene
```{r}

hg38.genes.bed.pos.mid.windows.all <- read.table("hg38.genes.pos.mid.windows.60w.bed",header=F) %>% 
                                  mutate(V4=V4+20) 

hg38.genes.bed.neg.mid.windows.all <- read.table("hg38.genes.neg.mid.windows.60w.bed",header=F) %>% 
                                  mutate(V4=V4+20) 

hg38.genes.bed.pos.early.windows.all <- read.table("hg38.genes.pos.early.windows.20w.bed",header=F) 

hg38.genes.bed.neg.early.windows.all <- read.table("hg38.genes.neg.early.windows.20w.bed",header=F) %>% 
                                    mutate(V4=V4+80)  

hg38.genes.bed.pos.late.windows.all <- read.table("hg38.genes.pos.late.windows.20w.bed",header=F) %>%
                                   mutate(V4=V4+80) 

hg38.genes.bed.neg.late.windows.all <- read.table("hg38.genes.neg.late.windows.20w.bed",header=F) 

# Combine the corrected bed files 
hg38.genes.corrected.all.bed <- rbind(hg38.genes.bed.pos.early.windows.all, hg38.genes.bed.pos.mid.windows.all, hg38.genes.bed.pos.late.windows.all,hg38.genes.bed.neg.early.windows.all, hg38.genes.bed.neg.mid.windows.all, hg38.genes.bed.neg.late.windows.all)
hg38.genes.pos.corrected.all.bed <- rbind(hg38.genes.bed.pos.early.windows.all, hg38.genes.bed.pos.mid.windows.all, hg38.genes.bed.pos.late.windows.all)
hg38.genes.neg.corrected.all.bed <- rbind(hg38.genes.bed.neg.early.windows.all, hg38.genes.bed.neg.mid.windows.all, hg38.genes.bed.neg.late.windows.all)

# Save files
write.table(hg38.genes.corrected.all.bed, "hg38.transcripts.corrected.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.genes.pos.corrected.all.bed, "hg38.transcripts.pos.corrected.bed", sep = "\t", quote = F, col.names = F, row.names = F)
write.table(hg38.genes.neg.corrected.all.bed, "hg38.transcripts.neg.corrected.bed", sep = "\t", quote = F, col.names = F, row.names = F)
```


---
Spike-in normalised metagene plot
---
lot the metagene plot using spike-in normalisation factor bigwigs
Once completed the mapping of the coverage onto the binned genome regions (in bed format), generate the plot (code from Ivan)


```{r}


### Plot 

# Define sample names and corresponding file prefixes
samples <- c("0h", "90min", "8h")
prefixes <- c("0h", "90min", "8h")

# Initialize empty list to store plot data frames
plot_list_si <- list()

for(i in seq_along(samples)) {
  sample_name <- samples[i]
  prefix <- prefixes[i]
  
  # Read forward and reverse coverage files and combine
  coverage <- rbind(
    read.table(paste0(prefix, "_spikein.norm._merged_hg38_fwd.clean.cov.bed"), header = FALSE),
    read.table(paste0(prefix, "_spikein.norm._merged_hg38_rev.clean.cov.bed"), header = FALSE)
  )
  
  # Rename columns
  colnames(coverage) <- c("chr", "start", "end", "bin", "COV")
  
  # Add dist column
  coverage <- coverage %>%
    mutate(dist = (as.numeric(bin) - 1) * 0.0505050505)
  
  # Summarise mean and SEM by dist
  plot_df <- coverage %>%
    group_by(dist) %>%
    summarise(
      COV.mean = mean(COV),
      SEM = sd(COV)/sqrt(n())
    ) %>%
    mutate(Condition = case_when(
      sample_name == "0h" ~ "0h",
      sample_name == "90min" ~ "90min",
      sample_name == "8h" ~ "8h"
    ))
  
  # Store result in list
  plot_list_si[[sample_name]] <- plot_df
}

# Combine all plot data frames
Total_genes.plot.si <- bind_rows(plot_list_si)
#only 0h and 90mins

Total_genes.plot.si_part <- Total_genes.plot.si %>% filter(Condition != "8h")
# Formatting for better aesthetics

Total_genes.plot.si$Condition <- factor(Total_genes.plot.si$Condition, 
                                     levels = c("0h", "90min", "8h"))

Total_genes.plot.si_part$Condition <- factor(Total_genes.plot.si_part$Condition, 
                                     levels = c("0h", "90min"))
#Plot the metagene plot
TT_metagene_si_p <- Total_genes.plot.si %>%
  ggplot(aes(x = dist, y = COV.mean, color = Condition, fill = Condition)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = COV.mean - SEM, ymax = COV.mean + SEM), alpha = 0.2, color = NA) +
  #scale_color_manual(values = colors) +
  #scale_fill_manual(values = colors) +
  #scale_linetype_manual(values = line_types) + # Add this to set linetypes
  scale_x_continuous(breaks = seq(0, 5, by = 1), 
                     labels = c("0" = "-5kb", "1" = "TSS", "2" = "33%", "3" = "66%", "4" = "TTS", "5" = "+5kb")) +
  scale_y_continuous(limits = c(0,25))+
  guides(linetype = guide_legend(override.aes = list(fill = NA))) + # Adjust the legend for linetypes
  xlab("Position") + ylab("Spike-in normalised coverage") + 
  ggtitle("") +
  theme_bw() +
  theme(
    #aspect.ratio = 1,
    legend.title = element_text(size = 14, face = "bold"), 
    legend.text = element_text(size = 10),
    axis.text.x = element_text(size = 11, angle = 0, vjust = 0, hjust = 0.5),
    axis.text.y = element_text(size = 11),
    axis.title.x = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title.y = element_text(size = 14, face = "bold", vjust = 2),
    plot.title = element_text(size = 16, face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    legend.background = element_blank(),
    legend.key = element_blank()
  ) +
  guides(fill = "none") + ggtitle("TT-seq metagene plot for non-overlapping all gene length (spike-in normalised)") # Hide the fill legend


#Figure 3F
TT_metagene_si_p_part <- Total_genes.plot.si_part %>%
  ggplot(aes(x = dist, y = COV.mean, color = Condition, fill = Condition)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = COV.mean - SEM, ymax = COV.mean + SEM), alpha = 0.2, color = NA) +
  #scale_color_manual(values = colors) +
  #scale_fill_manual(values = colors) +
  #scale_linetype_manual(values = line_types) + # Add this to set linetypes
  scale_x_continuous(breaks = seq(0, 5, by = 1), 
                     labels = c("0" = "-5kb", "1" = "TSS", "2" = "33%", "3" = "66%", "4" = "TTS", "5" = "+5kb")) +
  scale_y_continuous(limits = c(0,25), breaks= c(0, 10, 20))+
  guides(linetype = guide_legend(override.aes = list(fill = NA))) + # Adjust the legend for linetypes
  xlab("Position") + ylab("Spike-in normalised coverage") + 
  ggtitle("") +
  scale_color_manual(values = c( "0h" = "gray24",
    "90min" = "dodgerblue3",
    "Input" = "darkgrey"
  )) +
  scale_fill_manual(values = c(
     "0h" = "gray24",
    "90min" = "dodgerblue3",
    "Input" = "darkgrey"
  )) +
 theme(text = element_text(family = "sans"),
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.text.x = element_text(size = 14, color = "black"),
    axis.text.y = element_text(size = 14, color = "black"),
     axis.title.x = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    strip.text.x = element_text(size = 14),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    panel.background = element_rect(fill = "white", colour = NA)
  ) +
  guides(fill = "none")  # Hide the fill legend

ggsave("Figure3F.pdf",
       plot = TT_metagene_si_p_part,
       width = 4.5,
       height = 3.5,
       units = "in")
```

---
Extended analysis: further filtering by expression levels
---
compute baseMean and RPKM per group for data inspection
```{r}
#extract basemean

# get normalized counts
# normalized counts from your dds
norm_counts <- counts(dds_filtered_Degron, normalized=TRUE) %>%as.data.frame()

# get the condition info for each column
sample_info <- colData(dds_filtered_Degron)

# calculate group-wise mean normalized counts

baseMean_by_group <- 
  norm_counts %>%
  tibble::rownames_to_column("gene") %>%
  tidyr::pivot_longer(-gene, names_to="sample", values_to="norm_count") %>%
  left_join(data.frame(sample=rownames(sample_info),
                       condition=sample_info$group),
            by="sample") %>%
  group_by(gene, condition) %>%
  summarise(baseMean=mean(norm_count), .groups="drop") %>%
  tidyr::pivot_wider(names_from=condition, values_from=baseMean) %>% 
  rename(`0h` ="0h_avg_baseMean",  `8h` = "8h_avg_baseMean", `90min` = "90min_avg_baseMean")

#> summary(baseMean_by_group)
     #gene                 0h                  8h                90min          
 #Length:30506       Min.   :     0.00   Min.   :     0.00   Min.   :     0.00  
 #Class :character   1st Qu.:     4.64   1st Qu.:     5.40   1st Qu.:     4.73  
 #Mode  :character   Median :    34.19   Median :    41.91   Median :    36.28  
                    #Mean   :   949.38   Mean   :  1138.81   Mean   :   948.95  
                    #3rd Qu.:   559.27   3rd Qu.:   655.14   3rd Qu.:   564.50  
                    #Max.   :104236.74   Max.   :135689.65   Max.   :105605.61  

##get RPKM by Group

counts <- counts(dds_filtered_Degron) 

# total library size per sample
lib_sizes <- colSums(counts)

# keep only rows that are genes
genes_gtf <- gtf.dat[gtf.dat$type == "gene"]
# width gives the length of each gene
gene_lengths <- width(genes_gtf)

# assign gene IDs as names
names(gene_lengths) <- genes_gtf$gene_id

# match your count rownames (gene IDs) to gene lengths
gene_lengths_matched <- gene_lengths[rownames(counts)]

#use sweep to preserve gene names
rpkm <- sweep(counts, 2, lib_sizes, FUN = function(x, lib) x * 1e9 / (gene_lengths_matched * lib)) %>% 
        as.data.frame() 


sample_info_df <- sample_info %>% as.data.frame()


#generate RPKM by group
rpkm_by_group <- 
  rpkm %>%
  rownames_to_column("gene") %>%
  pivot_longer(-gene, names_to="sample", values_to="RPKM") %>%
  left_join(
    sample_info_df %>%
      dplyr::select(sample, group), 
    by="sample"
  ) %>%
  group_by(gene, group) %>%
  summarise(meanRPKM=mean(RPKM), .groups="drop") %>%
  pivot_wider(names_from=group, values_from=meanRPKM) %>% 
  rename(`0h` ="0h_avg_RPKM",  `8h` = "8h_avg_RPKM", `90min` = "90min_avg_RPKM")

#> summary(rpkm_by_group)
#     gene                 0h                  8h                90min          
# Length:30506       Min.   :   0.0000   Min.   :   0.0000   Min.   :   0.0000  
# Class :character   1st Qu.:   0.0103   1st Qu.:   0.0105   1st Qu.:   0.0108  
# Mode  :character   Median :   0.1167   Median :   0.1196   Median :   0.1212  
                    #Mean   :   0.9300   Mean   :   1.0008   Mean   :   0.9495  
                    #3rd Qu.:   0.7869   3rd Qu.:   0.7723   3rd Qu.:   0.7862  
                    #Max.   :1066.7297   Max.   :1244.8831   Max.   :1120.2572

```
Combine the RPKM and baseMeantable together

```{r}
#clean up the genes data
genes_gtf_clean <- genes_gtf %>% as.data.frame() %>% rename("gene_id" = "gene") %>% select("gene", "gene_name", "gene_biotype")

#combine the data together with gene name and gene biotype information
final_data <- left_join(rpkm_by_group,baseMean_by_group, by = "gene") %>% 
  left_join(genes_gtf_clean, by = "gene") %>% 
  select("gene", "gene_name", "gene_biotype", `0h_avg_RPKM` ,`8h_avg_RPKM`, `90min_avg_RPKM` ,`0h_avg_baseMean` ,`8h_avg_baseMean` ,`90min_avg_baseMean`)

write_csv(final_data, "Degron_Scott_genes_RPKM_baseMean.csv")
```


Generate the boxplot for basemean

```{r}
#convert into long format, then calculate the log2baseMean
log2baseMean_long <- baseMean_by_group %>% 
                      pivot_longer(!gene, names_to = "Condition", values_to = "baseMean") %>% 
                      mutate(log2baseMean = log2(baseMean)) %>% 
                      mutate(Condition = str_extract(Condition, "^[^_]+")) #clean up the suffixes

#reordering the levels of the conditions
log2baseMean_long$Condition <- factor(log2baseMean_long$Condition, 
                                     levels = c("0h", "90min", "8h"))
#plot the log2basemean boxplot
baseMean_boxplot <- ggplot(log2baseMean_long, aes(x = Condition, y = log2baseMean, fill = Condition)) +
  geom_boxplot(outlier.shape = NA) +
  theme_minimal() +
  #coord_cartesian(ylim = c(0,2)) +
  labs(x = "", y = "log2baseMean") +
  theme(
    legend.position = "right",
    legend.direction = "vertical",
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.text.x = element_blank(),
    axis.text.y = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    strip.text.x = element_text(size = 12),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 12),
    axis.ticks.y = element_blank(),
    panel.border = element_rect(color = "black", fill = NA, size = 1)
  ) 

```

---
Differential expression analysis
---

Perform differential expression analysis


No differential expression of genes observed. However, interestingly we do see the general shift of transcription to be lower ie closer to 0 in the 90min vs 0h in MA plot, vs 8h vs 0h MA plot
```{r}

res <- DESeq(dds_filtered_Degron)
#perform lfcshirnkage to reduce noise (cannot use alpgem as in Jespers paper as the model struggle t5o fiy)
resLFC_8hvs0h <- lfcShrink(res, coef="group_8h_vs_0h", type="ashr")

#> summary(resLFC_8hvs0h)

#out of 30506 with nonzero total read count
#adjusted p-value < 0.1
#LFC > 0 (up)       : 0, 0%
#LFC < 0 (down)     : 0, 0%
#outliers [1]       : 2, 0.0066%
#low counts [2]     : 0, 0%
#(mean count < 1)
#[1] see 'cooksCutoff' argument of ?results
#[2] see 'independentFiltering' argument of ?results
#[1] "Intercept"         "group_8h_vs_0h"    "group_90min_vs_0h"

time8hvs0h_res <- results(res, name="group_8h_vs_0h")
time90minvs0h_res <- results(res, name="group_90min_vs_0h")
resLFC_90minvs0h <- lfcShrink(res, coef="group_90min_vs_0h", type="ashr")
#> summary(time8hvs0h_res)
#out of 30506 with nonzero total read count
#adjusted p-value < 0.1
#LFC > 0 (up)       : 0, 0%
#LFC < 0 (down)     : 0, 0%
#outliers [1]       : 2, 0.0066%
#low counts [2]     : 0, 0%
#(mean count < 1)
#[1] see 'cooksCutoff' argument of ?results
#[2] see 'independentFiltering' argument of ?results


#> summary(resLFC_90minvs0h)

#out of 30506 with nonzero total read count
#adjusted p-value < 0.1
#LFC > 0 (up)       : 0, 0%
#LFC < 0 (down)     : 0, 0%
#outliers [1]       : 2, 0.0066%
#low counts [2]     : 0, 0%
#(mean count < 1)
#[1] see 'cooksCutoff' argument of ?results
#[2] see 'independentFiltering' argument of ?results
#> summary(time90minvs0h_res)

#out of 30506 with nonzero total read count
#adjusted p-value < 0.1
#LFC > 0 (up)       : 0, 0%
#LFC < 0 (down)     : 0, 0%
#outliers [1]       : 2, 0.0066%
#low counts [2]     : 0, 0%
#(mean count < 1)
#[1] see 'cooksCutoff' argument of ?results
#[2] see 'independentFiltering' argument of ?results

```


plot volcano plot
```{r}
time90minvs0h_res
# Prepare data
time90minvs0h_res_df <- time90minvs0h_res %>%
  as.data.frame() %>%
  mutate(
    neglog10p = -log10(pvalue),
    Status = ifelse(!is.na(padj) & padj < 0.05,
                         "Significant",
                         "Not Significant")
  )


# Volcano plot
ggplot(time90minvs0h_res_df , aes(x = log2FoldChange, y = neglog10p)) +
  geom_point(aes(color = Status), alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c("grey", "red")) +
  theme_minimal() +
  geom_vline(xintercept = c(-1.5, 1.5),
             linetype = "dotted",
             linewidth = 0.7) +
  labs(
    x = "log2 Fold Change (90min vs 0h)",
    y = "-log10(p-value)",
    title = "Volcano Plot: 90min vs 0h (by padj cut-off 0.05)"
  ) 


#export the data as csv
gene_names_df <- gtf.dat %>% as_tibble() %>% select(gene_id, gene_name) %>% distinct_all()

time90minvs0h_res_df2 <- time90minvs0h_res_df %>% rownames_to_column("gene_id") %>% select(-fdrtool_qval) %>% left_join(gene_names_df, by = "gene_id") %>% select(gene_id, gene_name, everything())

write_csv(time90minvs0h_res_df2, "DeSeq2_results_90minvs0h.csv")
```


Generate the MA plot overlaying with density map 
```{r}
plotMA(time8hvs0h_res, ylim=c(-2,2)) + title ("8h vs 0h MA plot")
plotMA(time90minvs0h_res, ylim=c(-2,2))+ title ("90min vs 0h MA plot")


#include the density plot to show where most of the reads are concentrated

library(MASS)      # for kde2d
library(ggplot2)
library(ggdag)
library(patchwork)


#time8hvs0h_res
# prepare data
res_df <- as.data.frame(time8hvs0h_res)
res_df <- res_df[!is.na(res_df$log2FoldChange) & !is.na(res_df$baseMean), ]
res_df$logBaseMean <- log10(res_df$baseMean + 1)

# compute 2D density
dens <- MASS::kde2d(res_df$logBaseMean, res_df$log2FoldChange, n = 200)
MAplot_density_8hvs0h <- ggplot(res_df, aes(x = logBaseMean, y = log2FoldChange)) +
  geom_point(alpha = 0.3, size = 0.8, color = "grey") +
  stat_density_2d(aes(fill = ..level.., alpha = ..level..), geom = "polygon", contour = TRUE,bins = 20) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "black") +
  scale_fill_viridis_c(option = "magma", name = "Density") +
  scale_alpha(range = c(0.2, 1.0), guide = "none") +
  scale_y_continuous(limits = c(-2, 2)) +
  theme_minimal() +
  labs(x = "log10(baseMean + 1)", y = "log2 Fold Change", 
       title = "8h vs 0h MA plot with density") + remove_grid()

#time90minvs0h_res

# prepare data
res_df2 <- as.data.frame(time90minvs0h_res)
res_df2 <- res_df2[!is.na(res_df2$log2FoldChange) & !is.na(res_df2$baseMean), ]
res_df2$logBaseMean <- log10(res_df2$baseMean + 1)

# compute 2D density (Figure 3E)
dens2 <- MASS::kde2d(res_df2$logBaseMean, res_df2$log2FoldChange, n = 200)
MAplot_density_90minvs0h <- ggplot(res_df2, aes(x = logBaseMean, y = log2FoldChange)) +
  geom_point(alpha = 0.3, size = 0.8, color = "grey") +
  stat_density_2d(aes(fill = ..level.., alpha = ..level..), geom = "polygon", contour = TRUE,bins = 20) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "black") +
  scale_fill_viridis_c(option = "magma", name = "Density") +
  scale_alpha(range = c(0.2, 0.8), guide = "none") +
  scale_y_continuous(limits = c(-2, 2)) +
  theme_minimal() +
  labs(x = "log10(mean expression)", y = "log2 Fold Change") + 
  theme(text = element_text(family = "sans"),
    legend.position = "right",
    legend.text = element_text(size = 14, color = "black"),
    axis.text.x = element_text(size = 14, color = "black"),
    axis.text.y = element_text(size = 14, color = "black"),
     axis.title.x = element_text(size = 14, color = "black"),
    axis.title.y = element_text(size = 14, color = "black"),
    strip.text.x = element_text(size = 14),
    axis.ticks = element_line(colour = "black"),
    axis.ticks.length = unit(0.2, "cm"),
    panel.border = element_rect(color = "black", fill = NA, size = 1),
    panel.background = element_rect(fill = "white", colour = NA)
  ) +remove_grid()


ggsave("Figure3E.pdf",
       plot = MAplot_density_90minvs0h,
       width = 4.5,
       height = 3.5,
       units = "in")
#use patchwork to combine the plots together
combined_MAdensity_plot <- MAplot_density_90minvs0h + MAplot_density_8hvs0h 
```

for the MA plots
If log10(baseMean + 1) ≈ 3.5, then

baseMean ≈ 3161




